# Covid19


```{r}
library(tidyverse)
library(haven)
library(gt)
library(gtsummary)
library(plotly)

# Function to get the name of the latest file
get_latest <- function(file_pattern){
    file_list <- list.files(path = "./data", pattern = file_pattern)
    latest_file <- tail(sort(file_list), n = 1)
    return(latest_file)
}


# Read file in the root of sub-folder "data" with name containing "ESS" 
df_ess <- read_dta(paste0("data/", get_latest("ESS")))

```


As shown in @fig-ess-time, the ESS data indicates a slightly higher average support for redistribution after the 2008 economic crisis and after the 2022 Covid19 pandemic. Evidence from a survey experiment in the USA show that participants are more willing to prioritize society's problems when exposed to issues related to the pandemic [@cappelen_solidarity_2021]. Van Hootegem and Laenen [-@hootegem_wave_2023] show that the increase in support for a universal basic income increased with the pandemic but it was short lived. 

In this section we explore further  the potential impact of the pandemic on support for redistribution. 


## Time series

@tbl-multiv-gincdif suggests that both social class and feeling about household income are significant predictors of support for redistribution. @fig-hincfel-time shows that the difference between subjective income groups remains stable over time, with all groups responding similar to 2008 crisis and the pandemic. 

```{r}
#| label: fig-hincfel-time
#| fig-cap: "Average support for redistribution by financial situation (2002-2022)"


get_series <- function(myvar){

(df_ess |>
    filter(!is.na(get(myvar))) |> 
    group_by(round = as_factor(essround), class = as_factor(get(myvar)))  |> 
    mutate(gincdif_inv = 6 - gincdif) |> 
    summarise(
        mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T),
        se = sd(gincdif_inv, na.rm = T)/sqrt(length((gincdif_inv)))) |> 
    ggplot(aes(x = round, y = mean, ymin = mean - 1.96*se, ymax = mean + 1.96*se, group = class, color = class)) +
    geom_pointrange() +
    geom_line()+
    ylim(3,4.5)+
    theme_classic()) |> 
    ggplotly()  |> 
    layout(legend = list(orientation = "h", x = 0.4, y = -0.2))
}

get_series("hincfel")

```

Conversely, the @fig-class5-time suggests that there is a slightly different pattern in response to the pandemic. There is a trajectory of increasing support for redistribution across skilled and unskilled workers since round 8 (2016). However, for lower-grade service class and small business owners, there seems to be a break in this trajectory captured in the latest wave. 

```{r}
#| label: fig-class5-time
#| fig-cap: "Average support for redistribution by social class (2002-2022)"

get_series("class5")

```


## Job loss


```{r}
#| label: fig-jobloss
#| fig-cap: "Average support for redistribution by social class (2002-2022)"

df_ess |> 
    filter(!is.na(class5) & !is.na(hapnoc19)) |> 
    mutate(
        impact = case_when(
            hapljc19 == 1 | hapfoc19 == 1 | hapfuc19 == 1 ~ "Lost job/unpaid leave/Forl.",
            hapirc19 == 1 | hapwrc19 == 1 ~ "Income / hours reduced",
            hapnoc19 == 1 ~ "Nothing",
            is.na(hapnoc19) ~ NA_character_,
            .default = "Other"
        )
    ) |> 
    group_by(impact, class = as_factor(class5)) |> 
    summarise(
        mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T),
        se = sd(gincdif_inv, na.rm = T)/sqrt(length((gincdif_inv)))) |> 
    ggplot(aes(x = class, y = impact, fill = mean))+
    geom_tile() +
    scale_fill_distiller(palette = "RdBu", direction=1) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

    # ggplot(aes(x = class, y = mean, ymin = mean - 1.96*se, ymax = mean + 1.96*se)) +
    # geom_pointrange() +
    # coord_flip()+
    # facet_wrap(~impact, nrow =1) 

```