# Annex: Data processing {.unnumbered}

## European Social Survey (ESS)

### Importing

Data downloaded from the European Social Survey [website](https://www.europeansocialsurvey.org) using the data builder.

The following file contain the cumulative dataset for Ireland across the 11 rounds available. A subset of relevant variables were selected before download.

```{r}
library(tidyverse)
library(haven)

ess_ie <- read_dta("./data/ess/ess_ie.dta")

```


### Invert scale of main outcome

Inverting scale of main outcome variable to denote stronger agreement with higher values. 

```{r}

ess_ie <- ess_ie |> mutate(gincdif_inv = 6 - gincdif)

```

### Year / quarter

Adding column identifying the year and the quarter in which the interview has started. 

```{r}
get_quarter <- function(month){
    if_else(is.na(month), NA,
        if_else(month < 4, 1,
            if_else(month > 3 & month < 7, 2,
                if_else(month > 6 & month < 10, 3, 
                    if_else(month> 9 & month < 13, 4, NA)))))
}

ess_ie <- 
    ess_ie |> 
        mutate(
            interv_quarter = case_when(     
                essround < 3 ~ paste0(inwyr,  get_quarter(inwmm)),
                essround > 2 & essround  < 10 ~ paste0(inwyys, get_quarter(inwmms)),
                essround > 9 ~ format(as.Date(inwds), "%Y%QQ")
            ),
            interv_year = case_when(
                essround < 3 ~ inwyr, 
                essround > 2 & essround  < 10 ~ inwyys,
                essround > 9 ~ as.numeric(format(as.Date(inwds), "%Y"))
            )
        )

```

### Social class

Create [Oesch social class schema](https://people.unil.ch/danieloesch/scripts/) from relevant columns in the dataset. 

```{r}
source("data/ess/Oesch_SocialClass.R")
```

### Exporting

Exporting dataset with minimal set of variables.

```{r}
ess_vars <- c(
    "cntry",
    "idno",
    "essround",
    "pspwght",
    "anweight",
    "agea",
    "gndr",
    "gincdif",
    "lrscale",
    "ctzcntr",
    "brncntr",
    "yrbrn",
    "isco08",
    "isco08p",
    "emplrel",
    "emplno",
    "emprelp",
    "iscoco",
    "iscocop",
    "emplnop",
    "mnactic",
    "eduyrs",
    "eisced",
    "edulvla",
    "hincfel",
    "hinctnta",
    "gincdif_inv",
    "interv_quarter",
    "interv_year",
    "class5",
    "class8",
    "class16",
    "brnocnt",
    "nmnybsc",
    "smdfslv",
    "dfincac",
    "gvcldcr",
    "gvslvol",
    "gvslvue",
    "imsclbn",
    "sbeqsoc",
    "sblazy",
    "sblwcoa",
    "sbprvpv",
    "sbstrec",
    "uemplwk",
    "gvhanc19",
    "respc19",
    "hapljc19",
    "hapfuc19",
    "hapfoc19",
    "hapirc19",
    "hapwrc19",
    "hapnoc19"
)

ess_ie <- ess_ie |> select(all_of(ess_vars))

write_dta(ess_ie, paste0("data/ESS_v", format(Sys.Date(), "%m%d"),".dta"))

```


## Voter Study (VS - European Elections Survey)

### Importing

Data downloaded from the GESIS [website](https://www.gesis.org/en/services/finding-and-accessing-data/international-survey-programs/european-election-studies).

```{r}
# Import individual files

vs_14 <- read_dta("./data/vs/ZA5160_v4-1-0.dta")
vs_19 <- read_dta("./data/vs/ZA7581_v2-0-1.dta")
vs_24 <- read_dta ("./data/vs/ZA8868_v1-0-0.dta")
```

### Combining files 

The surveys from 2014, 2019 and 2024 are available in three different files. The following code renames the main columns to 


```{r}

vs_vars <- c("cntry", "idno", "agea", "gndr", "redist", "weight", "study")

vs_14 <- 
    vs_14 |>
    mutate(
        study = "EES14",
        cntry = b,
        idno = respid,
        agea = vd11,
        gndr = as_factor(d10),
        redist = qpp17_2 -1, # see coding (0 on quest is = to 1 in dataset)
        weight = wex
    ) |> 
    select(all_of(vs_vars)) 

vs_19 <- 
    vs_19 |> 
    mutate(
        study = "EES19", 
        agea = 2019 - D4_1,
        cntry = hCountry,
        idno = respid,
        gndr = as_factor(D3),
        redist = Q14_2,
        weight = WGT1
    
    ) |> 
    select(all_of(vs_vars)) 

vs_24 <- 
    vs_24 |> 
    mutate(
        study = "EES24",
        cntry = country,
        idno = resp_id,
        agea = d4_age,
        gndr = as_factor(d3),
        redist = q12_2,
        weight = Weight2
    
    ) |> 
    select(all_of(vs_vars)) 


```

The following code combines these three different files into one cumulative dataset.

```{r}

vs_all <- bind_rows(vs_14, vs_19, vs_24)

```




### Missing values for main outcome

Assigning common NA code to all non-valid observations. 

```{r}
vs_all$redist <- if_else(vs_all$redist < -1 | vs_all$redist > 11, NA, vs_all$redist)

```


### Invert scale of main outcome

Assigning common NA code to all non-valid observations. 

```{r}
vs_all$redist_inv <- 10 - vs_all$redist
vs_all |>  count(redist, redist_inv)

```

### Select Ireland

Creating dataset with observation from Ireland only. 

```{r}
vs_ie  <- vs_all |> filter(cntry == 8)
```


### Exporting


```{r}
write_dta(vs_ie, paste0("data/VS_v", format(Sys.Date(), "%m%d"),".dta"))

```