---
title: "Public support for welfare and redistributive policies in Ireland"
format:
   docx:
    toc: false
    number-sections: true
    link-citations: true
    reference-doc: _assets/template.docx
fig-cap-location: bottom
#    pdf:
#     papersize: a4paper
#     documentclass: article
#     link-citations: true
#     fig-cap-location: top
#     fig-pos: "H"
execute: 
  echo: false
  message: false
  warning: false
  #fig-width: 2
# author:
#   - name: Daniel Capistrano
#     affiliations:
#       - ref: ESRI
#     orcid: 0000-0002-8157-5564
#     email: daniel.capistrano@esri.ie
#     url: https://danielcapistrano.com/

# abstract: |
#   In this study, we assessed .

# keywords: 
#   - Welfare
bibliography: _assets/CFI_Redistribution.bib
---

<style>
  .narrow {
    width: 60% !important;
  }
</style>

# Introduction

Many vulnerable individuals and families in Ireland, each with different profiles, circumstances, and needs, are struggling to make ends meet and are falling into poverty. A significant number of these people rely heavily on welfare benefits, but the current level of support is often insufficient to protect them from poverty. 

This report presents the findings of a research project that explored attitudes towards welfare and redistributive policies designed to address these challenges. Ireland offers a particularly relevant context for this research. Over the past few decades, the Irish society experienced different levels of economic growth, income inequality, and poverty @roantree_poverty_2024. Therefore, these changing conditions provide information to examine the factors associated with public support for welfare policies and income redistribution. 

The research project was guide by the following general questions:
- Who in Ireland supports welfare policies and income redistribution?  
- Has the overall support for welfare policies and income redistribution changed over time?  
- What are the main policy implications of this public support?  


## Key points

- Over time, there is an overall stable support for government's role  in reducing differences in income levels (@fig-ess-time) and for wealth redistribution (@fig-vs-time).

- ESS respondents from lower social classes and in worse financial situation are more likely to agree that the government should reduce differences in income levels (@tbl-multiv-gincdif).  

- From 2008 to 2016, there was a significant reduction on the average agreement that social benefits pplace too great strain on economy (@fig-benefits)  

- Respondents in worse financial situation (subjective self-reported) tend to disagree more with the perspective that social benefits are disincentives for work, economy and social cooperation (@tbl-multiv-beliefs)  


- There is a trajectory of increasing support for redistribution across skilled and unskilled workers since round 8 (2016). However, for lower-grade service class and small business owners, there seems to be a break in this trajectory captured in the latest wave (@fig-class5-time)


## Methods and data

Social surveys are widely used as a research instrument to assess welfare attitudes and attitudes to redistribution [@steele_attitudes_2019; @svallfors_contested_2012]

- **European Elections Studies**: The research examined public opinion data from Voter Surveys conducted immediately after the European elections. Data from the latest three surveys (2014,	2019, 2024) were analysed here. These surveys are part of the European Elections Studies (EES), which also features political elite surveys, media studies, and the Euromanifestos project that codes Party Manifestos. Before 1994, the voter survey questions 3 part of the Eurobarometer surveys. Since the 1999 elections, EES surveys have been conducted independently of the Eurobarometer. The main response variable used in this project is based on a question related to wealth redistribution. The interviewer manual does not contain any additional instruction in relation to this question, so respondents provided their answers according to their understanding of the following wording:

::: {.callout-note .narrow}

##### Question wording

[...] We would like  to ask you to position yourself on a scale from 0 to 10, where '0' means that you "fully agree with the statement at the top" and '10' means that you "fully agree with the statement at the bottom". [...]

**Redistribution of wealth**  
**0** - You fully favour redistribution from the rich to the poor in Ireland  
**10** You fully oppose redistribution of wealth from the rich to the poor in Ireland  

:::

 The original scale of this variable was inverted for this analysis. So, in the results presented here, higher values indicate more in favour of redistribution.


Here are relevant variables from the latest three waves that were also included in the analysis. 

- Country of birth  
- Age  
- Gender  
- Educational attainment  


- **European Social Survey**: The main data source for this project is the European Social Survey (ESS). Most of the analysis is based on a core question that was asked across all 11 rounds since 2002. In addition, we also explore several questions from the module on welfare attitudes that was fielded in 2008 and 2016. The main response variable used in this project is based on a question regarding the role of the government in reducing income inequalities that was administered across all 11 rounds. The interviewer manual does not contain any additional instruction in relation to this question, so respondents provided their answers according to their understanding of the following wording: 

::: {.callout-note}

#### Question wording

Using this card, please say to what extent you agree or disagree with each of the following statements. **The government should take measures to reduce differences in income levels**

1 Agree strongly  
2 Agree  
3 Neither agree nor disagree  
4 Disagree  
5 Disagree strongly  
:::

The original scale of this variable was inverted for this analysis. So, in the results presented here, higher values denote stronger agreement with government reducing income differences.

Some recent studies have raised concerns in relation to the reliability of this question as a measurement of support for redistribution. Margalit and Raviv -@margalit_does_2024 argues that reduction in income differences is too abstract and respondents in general do not link it to redistributive measures. Similarly, it has been suggested that this question only captures a diffuse inclination to equality and that is why responses are inconsistent with voting preferences across countries @dallinger_ambivalence_2022. However, @breznau_moderating_2025 argue that the absence of a relationship between this measurement and other expected correlates (voting preferences / support for concrete policies) is due to the ommission of views on government (trust and perceptions of corruption) as a moderating variable. 

Here are relevant variables from the latest three waves that were also included in the analysis. 

- Placement on left right scale  
- Citizen of country  
- Born in country  
- Gender   
- Year of birth   
- Highest level of education  
- Feeling about household's income nowadays  

In addition, based on the class scheme developed by Oesch -@oesch_coming_2006, a "social class" variable was constructed based on occupational variables and following  @tawfik_script_2020 script.  


# Who supports redistribution?

```{r}
library(tidyverse)
library(haven)
library(gt)
library(gtsummary)

# Function to get the name of the latest file
get_latest <- function(file_pattern){
    file_list <- list.files(path = "./data", pattern = file_pattern)
    latest_file <- tail(sort(file_list), n = 1)
    return(latest_file)
}


# Function to save an image of the table/plot and print the image on the report
save_display  <- function(myobj, w=NULL, h=NULL){
    name_str  <- deparse(substitute(myobj))
    file_path  <- paste0("_assets/img/", name_str, '.png')

    if(str_detect(name_str, 'tbl')){
        gtsave(myobj, filename = file_path)
    } else if (str_detect(name_str, 'fig')) {
        if(missing(w)){
            ggsave(myobj, filename = file_path)
        } else {
            ggsave(myobj, filename = file_path, width = w, height = h)
        }
    } else {
        print("Name of the object should contain 'fig' or 'tbl'")
    }

    knitr::include_graphics(file_path)
}

# Read file in the root of sub-folder "data" with name containing "ESS" 
df_ess <- read_dta(paste0("data/", get_latest("ESS")))

# Read file in the root of sub-folder "data" with name containing "VS" 
df_vs <- read_dta(paste0("data/", get_latest("VS_")))

# Read file in the root of sub-folder "data" with name containing "VSEU" 
df_vseu <- read_dta(paste0("data/", get_latest("VSEU_")))

# Read country with cross country data 
df_esseu <- read_dta("./data/ess/ess_eu.dta")



```


## Gender

The @tbl-gender shows the mean values for the main response variable in each study in 2024 by gender. 

```{r}
#| label: tbl-gender
#| tbl-cap: "Mean values by gender and study in 2024"

tbl_gender <- 
    df_ess |> 
        filter(essround == 11) |> 
        group_by(Gender = as_factor(gndr), .drop=TRUE) |>
        summarise(ESS = weighted.mean(gincdif_inv, w = pspwght, na.rm = T))  |> 
        left_join(
            df_vs |> 
            filter(study == "EES24") |> 
            group_by(Gender = as_factor(gndr), .drop=TRUE) |> 
            summarise(VS = weighted.mean(redist_inv, w = weight, na.rm = T))
        ) |> 
        gt() |> 
        fmt_number(everything(), decimals = 1) |> 
        tab_options(table.width = px(300))



save_display(tbl_gender)

```



## Age

The relationship between age and the response variable seems to change across rounds. As shown in @fig-ess-age, early rounds suggest that older respondents support more the government reducing income differences, whereas the latest round (11) and round 8 seem to indicate the opposite. 


::: {.panel-tabset}

### ESS

```{r}
#| label: fig-ess-age
#| fig-cap: "Scatter plot for 'Government should reduce income differences' by age and round"

fig_ess_age <- 
    df_ess |> 
        group_by(essround, agea) |> 
        summarise(mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T)) |> 
        ggplot(aes(x = agea, y = mean)) +
        geom_point() +
        geom_smooth(method = 'lm') +
        ylim(3,5) +
        facet_wrap(~essround)


save_display(fig_ess_age, w = 7, h =6)

```


### ESS

```{r}
#| label: fig-ess-cohort
#| fig-cap: "Mean values for 'Government should reduce income differences' by false cohort and round"

df_ess |> 
    mutate(false_cohort = cut(yrbrn, breaks = c(1900, 1940, 1980, 2020), right = FALSE, labels = c("1900-1940", "1940-1980", "1980-"))) |>
    filter(!is.na(false_cohort)) |> 
    group_by(essround, false_cohort) |> 
    summarise(mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T)) |> 
    ggplot(aes(x = essround, y = mean, colour = false_cohort, fill = false_cohort)) +
    geom_line()+
    geom_point() +
    ylim(3,5) +
    theme_classic()

```


### VS

```{r}
#| label: fig-vs-age
#| fig-cap: "Scatter plot for 'wealth redistribution' by age and study"

df_vs |> 
    group_by(study, agea) |> 
    summarise(mean = weighted.mean(redist_inv, w = weight, na.rm = T)) |> 
    ggplot(aes(x = agea, y = mean)) +
    geom_point() +
    geom_smooth(method = 'lm') +
    facet_wrap(~study)


```

:::


## Social class

@fig-class5 suggests a negative relationship between social class and the main ESS response variable. The higher the social class, the lower the support for the government to reduce income differences. 

::: {.panel-tabset}

### 5 groups

```{r}
#| label: fig-class5
#| fig-cap: "Scatter plot for 'income differences' by social class (5 groups) and round"

df_ess |> 
    mutate(class_inv = 6 - class5) |> 
    group_by(essround, class = class_inv) |> 
    summarise(mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T)) |> 
    ggplot(aes(x = class, y = mean)) +
    geom_point() +
    geom_smooth(method = 'lm') +
    facet_wrap(~essround)


```


### 16 groups


```{r}
#| label: fig-class16
#| fig-cap: "Scatter plot for 'income differences' by social class (16 groups) and round"



df_ess |> 
    mutate(class_inv = 17 - class16) |> 
    group_by(essround, class = class_inv) |> 
    summarise(mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T)) |> 
    ggplot(aes(x = class, y = mean)) +
    geom_point() +
    geom_smooth(method = 'lm') +
    facet_wrap(~essround)

```



:::


## Multivariate

The @tbl-multiv-gincdif below shows the coefficients of an OLS regression model on the support for government reducing differences in income levels. The overall explanatory power of the model is low. The effect of age seems to be neglectable. Both the social class and subjective income suggest that respondents in more precarious economic situation tend to be more favourable of the government reducing income differences. 

```{r}
##| label: tbl-multiv-gincdif
##| tbl-cap: "OLS regression coefficients for 'income differences' "


lm(gincdif_inv ~ factor(essround) + agea + as_factor(gndr) + as_factor(class5) + as_factor(hincfel) + cso_unemp, weights = pspwght, data = df_ess)  |> 
    tbl_regression(
        include = c("agea", "as_factor(gndr)", "as_factor(class5)", "as_factor(hincfel)", "cso_unemp"),
    )  |> 
    bold_p(t = 0.05, q = FALSE) |> 
    add_glance_source_note() |> 
    as_gt() |>
    gt::tab_source_note("Time fixed effects (ESS Round) omitted from the table but included in the model") |> 
    tab_options(table.layout = "auto")


```


## Has the support changed over time?

The average level of support for income redistribution in Ireland is similar to the ones observed in other European countries. However, the average support varies substantially over time. The @fig-ess-time below shows the mean values for 'Government should reduce income differences' by round. Overall there is a stable agreement (mean = `r round(weighted.mean(df_ess$gincdif_inv, w = df_ess$pspwght, na.rm = T),1)`). Rounds 5-7 (2010-2014) as well as 10-11(2022-2024) show a slightly higher agreement compared to other rounds and the European average.  

```{r}
#| label: fig-ess-time
#| fig-cap: "Mean values for 'Government should reduce income differences' by round" 

list_countries <- c("IE", "GR", "PT", "GB", "BE", "CZ", "FI", "FR", "DE", "NO", "ES")


df_esseu |>
    filter(cntry %in% list_countries) |> 
    mutate(Country = case_match(cntry,
        "IE" ~ "Ireland",
        "PT" ~ "Portugal",
        # "GR" ~ "Greece",
        "GB" ~ "Great Britain",
        .default = paste0("Other ", length(list_countries), " EU countries"))) |> 
    group_by(round = as_factor(essround), Country)  |> 
    mutate(gincdif_inv = 6 - gincdif) |> 
    summarise(
        mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T),
        se = sd(gincdif_inv, na.rm = T)/sqrt(length((gincdif_inv)))) |> 
    ggplot(aes(x = round, y = mean, group = Country, color = Country)) +
    geom_line()+
    geom_point() +
    ylim(3,5) +
    theme_classic()

    
```

@fig-vs-time shows the density of responses for the scale on support for redistribution of wealth. The mean values are similar for 2014, 2019 and 2024. However, the distribution seems is less skewed in 2024 compared to 2014.

```{r}
#| label: fig-vs-time
#| fig-cap: "Density plot for 'in favour of redistribution of wealth' by survey year"

df_vseu |>
    mutate(
        Country = if_else(as_factor(cntry) == "IRELAND", "Ireland", paste0("Other EU countries")),
        Year = as.numeric(str_remove(study, "EES"))
    ) |> 
    group_by(Year, Country)  |> 
    summarise(
        mean = weighted.mean(redist_inv, w = weight, na.rm = T),
        se = sd(redist_inv, na.rm = T)/sqrt(length((redist_inv)))) |> 
    ggplot(aes(x = Year, y = mean, group = Country, color = Country)) +
    geom_line()+
    geom_point() +
    ylim(4,7) +
    theme_classic()


# df_vseu |> 
#     ggplot(aes(x = redist_inv)) + 
#     geom_density(fill = "lightblue") + 
#     geom_vline(data = df_vs |> group_by(study) |> summarise(mean = weighted.mean(redist_inv, w = weight, na.rm = T)), aes(xintercept = mean), color = 'red') +
#     facet_wrap(~study, nrow = 3) +
#     theme_classic()

```


# Economic crisis and Covid19

As shown in @fig-ess-time, the ESS data indicates a slightly higher average support for redistribution after the 2008 economic crisis and after the 2022 Covid19 pandemic. Evidence from a survey experiment in the USA show that participants are more willing to prioritize society's problems when exposed to issues related to the pandemic [@cappelen_solidarity_2021]. Van Hootegem and Laenen [-@van_hootegem_wave_2023] show that the increase in support for a universal basic income increased with the pandemic but it was short lived. 

In this section we explore further  the potential impact of the pandemic on support for redistribution. 


## Time series

@tbl-multiv-gincdif suggests that both social class and feeling about household income are significant predictors of support for redistribution. @fig-hincfel-time shows that the difference between subjective income groups remains stable over time, with all groups responding similar to 2008 crisis and the pandemic. 

```{r}
#| label: fig-hincfel-time
#| fig-cap: "Average support for redistribution by financial situation (2002-2022)"


get_series <- function(myvar){

df_ess |>
    filter(!is.na(get(myvar))) |> 
    group_by(round = as_factor(essround), class = as_factor(get(myvar)))  |> 
    mutate(gincdif_inv = 6 - gincdif) |> 
    summarise(
        mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T),
        se = sd(gincdif_inv, na.rm = T)/sqrt(length((gincdif_inv)))) |> 
    ggplot(aes(x = round, y = mean, ymin = mean - 1.96*se, ymax = mean + 1.96*se, group = class, color = class)) +
    geom_pointrange() +
    geom_line()+
    ylim(3,4.5)+
    theme_classic()
}

get_series("hincfel")

```

Conversely, the @fig-class5-time suggests that there is a slightly different pattern in response to the pandemic. There is a trajectory of increasing support for redistribution across skilled and unskilled workers since round 8 (2016). However, for lower-grade service class and small business owners, there seems to be a break in this trajectory captured in the latest wave. 

```{r}
#| label: fig-class5-time
#| fig-cap: "Average support for redistribution by social class (2002-2022)"

get_series("class5")

```


## Job loss


```{r}
#| label: fig-jobloss
#| fig-cap: "Average support for redistribution by social class (2002-2022)"

df_ess |> 
    filter(!is.na(class5) & !is.na(hapnoc19)) |> 
    mutate(
        impact = case_when(
            hapljc19 == 1 | hapfoc19 == 1 | hapfuc19 == 1 ~ "Lost job/unpaid leave/Forl.",
            hapirc19 == 1 | hapwrc19 == 1 ~ "Income / hours reduced",
            hapnoc19 == 1 ~ "Nothing",
            is.na(hapnoc19) ~ NA_character_,
            .default = "Other"
        )
    ) |> 
    group_by(impact, class = as_factor(class5)) |> 
    summarise(
        mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T),
        se = sd(gincdif_inv, na.rm = T)/sqrt(length((gincdif_inv)))) |> 
    ggplot(aes(x = class, y = impact, fill = mean))+
    geom_tile() +
    scale_fill_distiller(palette = "RdBu", direction=1) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

    # ggplot(aes(x = class, y = mean, ymin = mean - 1.96*se, ymax = mean + 1.96*se)) +
    # geom_pointrange() +
    # coord_flip()+
    # facet_wrap(~impact, nrow =1) 

```


## Support for welfare policies

The data analysed in this chapter refers to a special module from the European Social Survey on welfare attitudes that was fielded in 2008 (Round 4) and repeated in 2016 (Round 8) (see @meuleman_past_2018 for more on this module). 


## Large differences acceptable

In this question, respondents were asked to what extent they agree with the following statement: 

::: {.callout-note collapse="true"}

### Question wording
‘Large differences in people’s incomes are acceptable to properly reward differences in talents and efforts’
:::

The distribution appears to be similar for 2008 and 2016, with a slight reduction in the proportion of those who "agree" or "agree strongly". 

```{r}
#| label: fig-large-diff
#| fig-cap: "Distribution of 'Large differences acceptable' (2008 and 2016)"

df_ess |> 
    filter(!is.na(dfincac)) |> 
    group_by(essround) |>
    count(Acceptable = as_factor(dfincac)) |> 
    mutate(Proportion = round(prop.table(n)*100,1)) |> 
    ggplot(aes(x = Acceptable, y = Proportion)) +
    geom_bar(stat = "identity", fill = "lightblue")+
    geom_label(aes(label = Proportion)) +
    theme_classic()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")+
    facet_wrap(~essround) 

df_ess |> 
 count(class5, class16)
```

A multivariate analysis with the same explanatory variables from the model  in @tbl-multiv-gincdif suggests that gender and class are not statistically significant predictors. However, respondents with worse financial situation tend to disagree more with this statement.

```{r}
#| include: false
lm(dfincac ~ factor(essround) + agea + as_factor(gndr) + as_factor(class5) + as_factor(hincfel), weights = pspwght, data = df_ess)  |> 
    tbl_regression() |> 
    bold_p(t = 0.05, q = FALSE)
```



## Deservingness

In this question, respondents were asked how much responsibility the government should have in relation to the following groups:

::: {.callout-note collapse="true"}

### Question wording

People have different views on what the responsibilities of governments should or should not be. For each of the tasks I read out please tell me on a score of 0-10 how much responsibility you think governments should have. 0 means it should not be governments' responsibility at all and 10 means it should be entirely governments' responsibility

"... ensure sufficient child care services for working parents"  

"... ensure a reasonable standard of living for the old" 

"...ensure a reasonable standard of living for the unemployed"
:::

In the @fig-deservingness, higher values denote higher government responsibility. On average respondents believe that the government should be more responsible for the old compared to working parents and unemployed. However, there was a reduction in this prioritisation for governments' responsibility in relation to older citizens and unemployed and increase in the average prioritisation for working parents.  

```{r}
#| label: fig-deservingness
#| fig-cap: "Mean values for deservingness of government support by group and round"


df_ess |>
    filter(essround %in% c(4, 8)) |> 
    group_by(round = as_factor(essround))  |> 
    summarise(across(c("gvcldcr", "gvslvol", "gvslvue"), ~ weighted.mean(.x, w = pspwght, na.rm = T))) |> 
    pivot_longer(cols = -round, names_to = "Group", values_to = "mean") |> 
    mutate(Group = case_match(Group, "gvcldcr" ~ "working parents", "gvslvol" ~ "old", "gvslvue" ~ "unemployed")) |> 
    ggplot(aes(x = round, y = mean, group = Group, color = Group)) +
    geom_point() +
    geom_line() +
    geom_label(aes(label = round(mean, 1)))+
    theme_classic() +
    theme(legend.position="bottom")

```

## Beliefs about social benefits

The module also included questions in relation to beliefs about the consequences of social benefits and services: 

::: {.callout-note collapse="true"}

### Question wording

Using this card please tell me to what extent you agree or disagree that social benefits and services in [country]... 

"...lead to a more equal society?"  

"... make people lazy?" 

"...make people less willing care for one another?"

"...prevent widespread poverty"  

"...place too great strain on economy"

:::

The average level of agreement remains similar comparing 2008 and 2016. The only exception is the average for "social benefits place too great strain on economy", which oberved a significant reduction in the level of agreement, as shown in @fig-benefits. 

```{r}
#| label: fig-benefits
#| fig-cap: "Mean values for beliefs in relation to social benefits by round"

believe_vars <- c("sbeqsoc", "sblazy", "sblwcoa", "sbprvpv", "sbstrec")

df_ess <- df_ess |> mutate(across(all_of(believe_vars), ~ 6 - .x, .names = "{.col}_inv"))

believe_vars_inv <- c("sbeqsoc_inv", "sblazy_inv", "sblwcoa_inv", "sbprvpv_inv", "sbstrec_inv")


df_ess |>
    filter(essround %in% c(4, 8)) |> 
    group_by(round = as_factor(essround))  |> 
    summarise(across(all_of(believe_vars_inv), ~ weighted.mean(.x, w = pspwght, na.rm = T))) |> 
    pivot_longer(cols = -round, names_to = "Question", values_to = "mean") |> 
    mutate(Question = case_match(
        Question, 
        "sbeqsoc_inv" ~ "Lead to equal society", 
        "sblazy_inv" ~ "Make people lazy", 
        "sblwcoa_inv" ~ "Make people less willing to care",
        "sbstrec_inv" ~ "Put strain on economy",
        "sbprvpv_inv" ~ "Prevent poverty"
    )) |> 
    ggplot(aes(x = round, y = mean, group = Question, color = Question)) +
    geom_point() +
    geom_line() +
    # geom_label(aes(label = round(mean, 1)))+
    theme_classic()
    
```


### PCA

A Principal Components Analysis indicate that these five variables can be reduced to the following two latent dimensions named arbitrarily:

- **Protective** (PC1): "Lead to equal society" and "Prevent poverty";    
- **Disincentive** (PC2): "Make people lazy" and "Make people less willing to care", and "Put strain on economy".

```{r}
#| include: false
library(psych)

pca_res <- 
    df_ess |> 
    select(all_of(believe_vars_inv)) |>  
    na.omit() |> 
    psych::principal(nfactors = 2)

pca_vars  <- data.frame(unclass(pca_res$loadings))

```

The variables were aggregated within each dimension using the mean value of the non-missing answers. The @tbl-multiv-beliefs below shows the coefficients of two OLS explanatory models for these two dimensions.

The first model indicates that female respondents tend to provide a slightly lower score on the protective dimension of beliefs about social benefits. None of the other predictors included in the model are statistically significant. 

The second model suggests that the group of "small business owners" and "skilled workers" have higher scores compared to respondents of other social classes. In addition, respondents with worse self-reported financial situation tend to disagree more with beliefs related to disincentive dimension of social beliefs. 

::: {.landscape}


```{r}
#| label: tbl-multiv-beliefs
#| tbl-cap: "OLS regression coefficients for 'beliefs about social benefits' "


vars_protect <- c("sbeqsoc_inv", "sbprvpv_inv")
vars_disinctv <- c("sblazy_inv", "sblwcoa_inv", "sbstrec_inv")

df_ess$protective <- rowMeans(select(df_ess, all_of(vars_protect)), na.rm = T)
df_ess$disincentive <- rowMeans(select(df_ess, all_of(vars_disinctv)), na.rm = T)

model_belief <- function(DV){
    model <- as.formula(paste0(DV, "~ factor(essround) + agea + as_factor(gndr) + as_factor(class5) + as_factor(hincfel)" ))

    lm(model, weights = pspwght, data = df_ess) |> 
        tbl_regression(include = c("agea", "as_factor(gndr)", "as_factor(class5)", "as_factor(hincfel)"))  |> 
        bold_p(t = 0.05, q = FALSE) 
}


tbl_merge(
    tbls = map(c("protective", "disincentive"), model_belief),
    tab_spanner = c("**Protective**", "**Disincentive**")) |> 
    as_gt() |>
    gt::tab_source_note("Time fixed effects (ESS Round) omitted from the table but included in the model") |> 
    tab_options(table.layout = "auto")



```

:::

# References {.unnumbered}

::: {#refs}
::: 