---
title: "Public support for welfare and redistributive policies in Ireland"
format:
   docx:
    toc: false
    number-sections: true
    link-citations: true
    reference-doc: _assets/esri_template.docx
    fig-cap-location: top
    prefer-html: true
knitr:
  opts_chunk:
    out.width: "70%"
execute: 
  echo: false
  message: false
  warning: false
bibliography: _assets/CFI_Redistribution.bib
---

# Introduction {#sec-introduction} 

Many vulnerable individuals and families in Ireland, each with different profiles, circumstances, and needs, are struggling to make ends meet and are falling into poverty. A significant number of these people rely heavily on welfare benefits, but the current level of support is often insufficient to protect them from poverty. In a context of increasing challenges to social welfare in Ireland, it is crucial to have a better understanding of the public support to State actions designed to address these challenges.  

With this objective, the current report presents findings of a research project that explored attitudes towards welfare and redistributive policies among Irish residents. Ireland offers a particularly relevant context for this research. Over the past few decades, the Irish society experienced different levels of economic growth, market inequality, and poverty [@roantree_poverty_2024]. In this sense, the changing conditions provide valuable information to examine the factors associated with public support for welfare policies and income redistribution. 

The research project was guide by the following general questions:  

- **Who** in Ireland supports income redistribution?  

- **What** Irish residents think about welfare policies?    

- Has the overall support for welfare policies and income redistribution **changed** over time?   

[Chapter @sec-who] in this report addresses the first question of a diffuse and generalised support for income and wealth redistribution in the Irish society. [Chapter @sec-what], addressing the second question, examines more specific support to welfare policies, beliefs about social benefits and their beneficiaries. Finally, [Chapter @sec-change] looks at change over time in relation to these topics and explores the possible impact of the 2008 economic recession and the 2020 pandemic.

## Key points

- ESS respondents from lower social classes and in worse financial situation are more likely to agree that the government should be responsible for reducing differences in income levels (@tbl-multiv-gincdif).  

- There is a slightly higher average support for income redistribution in Ireland compared to other northern European countries (@fig-eumap).  

- Ireland has one of the highest proportions in Europe of respondents who believe that 'social benefits make people lazy' (@fig-lazy).  

- Younger, right-wing, and those with lower educational attainment are more likely to agree that social benefits make people lazy (@tbl-multiv-beliefs).  

- From 2009 to 2016 there was an increase in the proportion of respondents who agree that 'social benefits lead to a more equal society' (@fig_benefits).    

- Tax policy has moved in the same direction as public opinion but not unemployment policy @rosset_unevenly_2025

- Baumberg Geiger: the main difference between the Scandinavian nations and the rest of Europe is not the extent to which negative attitudes to welfare exist, but the degree to which positive attitudes do

- Ernst Fehr: Inequality aversion, selfish and altruism in support for redistribution


## Methods and data

Social researchers make use of a varied set of methods to assess welfare attitudes. However, the main quantitative method in this field is social survey [@steele_attitudes_2019; @svallfors_contested_2012]. Despite their limited capacity to capture nuance and ambivalence in welfare attitudes [@goerres_can_2012], surveys help researchers to identify trends and patterns within and across different societies. 

In this sense, the main data source selected for this project is the European Social Survey (ESS)[^ess]. More than 20 European countries participate in this survey that is conducted roughly every two years. Ireland is one of the few countries that participated in all 11 rounds since 2002. On average, 2,200 people who were 18 or older and resident in Ireland were interviewed in each of these rounds. 

This research analysed several questions from a special module on welfare attitudes that was fielded in 2009 and 2016. However, most of the analysis is based on a core question, included in all rounds, on the role of the government in reducing income inequalities. Beyond that, the ESS questionnaire contains several other questions relevant to this research related to socio-demographic characteristics. In addition, based on the class scheme developed by @oesch_coming_2006, a "social class" variable was constructed using questions about the respondent's occupation (following @tawfik_script_2020 script).  

[^ess]: For more info, visit https://www.europeansocialsurvey.org 

Finally, this research also relied on public opinion data from Voter Surveys (VS) conducted immediately after the European elections (2014,	2019, 2024). These surveys are part of the European Elections Studies (EES), which also features political elite surveys, media studies, and the Euromanifestos project that codes Party Manifestos. Despite being part of the Eurobarometer project before 1994, thee EES surveys have been conducted independently since 1999. The survey offers additional information particularly on electoral behavious and political attitudes. 



# Who supports redistribution? {#sec-who} 

```{r}
#| include: false

library(tidyverse)
library(haven)
library(gt)
library(gtsummary)
library(rnaturalearth)
library(rnaturalearthdata)

# Function to get the name of the latest file
get_latest <- function(file_pattern){
    file_list <- list.files(path = "./data", pattern = file_pattern)
    latest_file <- tail(sort(file_list), n = 1)
    return(latest_file)
}


# Function to save an image of the table/plot and print the image on the report
save_display  <- function(myobj, w=NULL, h=NULL, vwidth = 600){
    name_str  <- deparse(substitute(myobj))
    file_path  <- paste0("_assets/img/", name_str, '.png')

    if(str_detect(name_str, 'tbl')){
        gtsave(myobj, filename = file_path, vwidth = vwidth)
    } else if (str_detect(name_str, 'fig')) {
        if(missing(w)){
            ggsave(myobj, filename = file_path, dpi=320)
        } else {
            ggsave(myobj, filename = file_path, width = w, height = h, dpi=320)
        }
    } else {
        print("Name of the object should contain 'fig' or 'tbl'")
    }
    
    knitr::include_graphics(file_path)
}

```

```{r}
#| include: false

# Read file in the root of sub-folder "data" with name containing "ESS" 
df_ess_raw <- read_dta(paste0("data/", get_latest("ESS")))

# Getting complete observations
df_ess <- df_ess_raw |> filter(!if_any(c(agea, gndr, hincfel), ~is.na(.x)))

# Read file in the root of sub-folder "data" with name containing "VS" 
df_vs <- read_dta(paste0("data/", get_latest("VS_")))

# Read file in the root of sub-folder "data" with name containing "VSEU" 
df_vseu <- read_dta(paste0("data/", get_latest("VSEU_")))

# Read country with cross country data 
df_esseu_raw <- read_dta("./data/ess/ess_eu.dta")

df_esseu <- df_esseu_raw |> filter(!if_any(c(agea, gndr, hincfel), ~is.na(.x)))

```


The ESS question used to assess support for redistribution asked participants to indicate if the government should take measures to reduce differences in income levels. The interviewer manual does not contain any additional instruction in relation to this question, so respondents provided their answers according to their understanding of the following wording: 

::: {.callout-note}

#### Question wording

Using this card, please say to what extent you agree or disagree with each of the following statements. **The government should take measures to reduce differences in income levels**.   

1 - Agree strongly   

2 - Agree   

3 - Neither agree nor disagree   

4 - Disagree   

5 - Disagree strongly   
:::

The original scale of this variable was inverted for our analysis. So, in the results presented in this report, higher values denote stronger agreement with government reducing income differences.  

Recent studies have raised concerns in relation to the reliability of this question as a measurement of support for redistribution. @margalit_does_2024 argues that reduction in income differences is too abstract and respondents in general do not link it to redistributive measures. Similarly, it has been suggested that this question only captures a diffuse inclination to equality and that is why responses are inconsistent with voting preferences across countries [@dallinger_ambivalence_2022]. However, @breznau_moderating_2025 argue that the absence of a relationship between this measurement and other expected correlates (voting preferences, support for concrete policies) is due to the omission of views on government (trust and perceptions of corruption) as a moderating variable. 

Similar to the main ESS question, the VS questionnaires also ask whether voters are in favour of wealth redistribution from the rich to the poor using a scale from 0 ('fully oppose') to 10 ('fully favour'). The original scale of this variable was inverted for this analysis. So, in the results presented here, higher values indicate more in favour of redistribution.

## Higher support in Ireland compared to other northern European countries

The mean levels of support for income redistribution in 2022 shown in @fig-eumap is in line with the south/north geographical divide identified before @meuleman_past_2018, with @finseraas_income_2009 arguing that countries with higher income inequality tend to be more supportive of income redistribution. The mean value for Ireland (3.95) is similar to the one observed in France (3.90) and slightly higher compared to other countries in the north of Europe such as the United Kingdom (3.82) and the Netherlands (3.63). 

```{r}
#| label: fig-eumap
#| fig-cap: "Mean value for 'Governement should reduce income differences', Europe, 2023"

eumap <- ne_countries(scale = "medium", continent = "Europe", returnclass = "sf") |> filter(!name %in% c("Russia", "Faeroe Is."))

fig_eumap <- 
    eumap |> 
        mutate(cntry = case_when(
            name == 'France' ~ "FR",
            name == "Norway" ~ "NO",
            .default = iso_a2)) |> 
        left_join(
            df_esseu |>
            filter(essround == 11) |> 
            group_by(cntry) |> 
            mutate(gincdif_inv = 6 - gincdif) |> 
            summarise(
                mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T),
                se = sd(gincdif_inv, na.rm = T)/sqrt(length((gincdif_inv)))),
            by = "cntry"
        ) |> 
        ggplot() +
        geom_sf(aes(fill = mean)) +
        scale_fill_distiller(palette = "RdYlBu", na.value = "white") +
        coord_sf(xlim = c(-25, 40), ylim = c(25, 75), expand = FALSE) +
        theme_void() +
        labs(
            fill = "Government should \nreduce income differences",
            caption = "Source: European Social Survey, Round 11"
        ) +
        theme(
            legend.position = "left", 
            legend.title = element_text(size = 8),
            plot.caption = element_text(hjust = -1)
        )  

save_display(fig_eumap, w =5, h=4)

```


## Female and lower social class respondents are more supportive of redistribution

Despite the profusion of studies on the relationship between gender and welfare policies, there is a gap in understanding the gender dimension of beliefs and perceptions in relation to these policies @garritzmann_gender_2021. Considering the marked gender differences in access to the labour market and levels of pay in Ireland @hingre_gender_2024, it is plausible to also expect differences in demands for income redistirbution. 

On average, it is not possible to identify statistically significant differences for male and female individuals in the latest rounds of both surveys. The ESS question on the government role has a similar mean values for male (3.9) and female (3.9) respondents. Similarly, the voter study identified similar mean values for wealth redistribution across male (5.4) and female (5.2) participants.

```{r}
#| include: false
df_ess |> 
    filter(essround == 11) |> 
    group_by(Gender = as_factor(gndr), .drop=TRUE) |>
    summarise(ESS = weighted.mean(gincdif_inv, w = pspwght, na.rm = T))  |> 
    left_join(
        df_vs |> 
        filter(study == "EES24") |> 
        group_by(Gender = as_factor(gndr), .drop=TRUE) |> 
        summarise(VS = weighted.mean(redist_inv, w = weight, na.rm = T))
    ) 

lm(gincdif_inv ~ as_factor(gndr) + essround, weights = pspwght, data= df_ess) |>  summary()

```

However, as shown in @fig-sex, there is some variation across time in relation to this gap. Considering all 11 rounds of the ESS together, there is a significantly higher support for redistribution among female respondents compared to male respondents. 

```{r}
#| label: fig-sex
#| fig-cap: "Mean value for 'Governement should reduce income differences' in Ireland by sex, 2002-2023"

fig_sex <- 
    df_ess |> 
        group_by(essround, Sex = as_factor(gndr), .drop=TRUE) |>
        summarise(
                mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T),
                se = sd(gincdif_inv, na.rm = T)/sqrt(length((gincdif_inv)))
            ) |> 
            ggplot(aes(x = essround, y = mean,  ymin = mean - 1.96*se, ymax= mean + 1.96*se, fill = Sex, colour = Sex)) +
            geom_pointrange() +
            geom_line() +
            theme_classic() +
            labs(
                caption = "Source: European Social Survey"
            ) +
            theme(
                legend.position = "bottom", 
                legend.title = element_text(size = 8),
                plot.caption = element_text(hjust = -1)
            )

save_display(fig_sex, w=6, h=4)

```

In addition, data suggest an age pattern where younger respondents tend to be more supportive of income redistribution in Ireland. As shown in @fig-age-eu, the opposite pattern is observed in Germany, Finland, France, and Portugal. Although the UK also has a similar negative relationshop between age and  support, it seems to be less pronounced compared to Ireland. 

```{r}
#| label: fig-age-eu
#| fig-cap: "Scatter plot for 'reduce income differences' by age and country, 2023"

fig_age_eu <- 
    df_esseu |> 
        mutate(gincdif_inv = 6 - gincdif) |> 
        filter(essround == 11 & cntry %in% c('IE', 'FR', 'FI', 'GB', 'NL', 'DE')) |> 
        group_by(Country = as_factor(cntry), agea) |> 
        summarise(mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T)) |> 
        ggplot(aes(x = agea, y = mean)) +
        geom_point(size = 1) +
        geom_smooth(method = 'lm') +
        ylim(3,5) +
        facet_wrap(~Country) + 
        labs(
            x = "Age", 
            y = "Government should reduce income differences", 
            caption = "Source: European Social Survey 2023"
        )+
        theme_classic() +
        theme(plot.caption = element_text(hjust = 0))

save_display(fig_age_eu, w=6, h=4)

```

However, zooming out for a longer time period, we observe different relationships depending on the time of the data collection. During the third and fourth rounds, there is a positive association between age and support for redistribution. This different associations over time might suggest a different level of support across generations instead of age groups. This topic is further explored in [Chapter @sec-change].

```{r}
#| label: fig-age-round
#| fig-cap: "Scatter plot for 'reduce income differences' by age and ESS round, Ireland"

fig_age_round <- 
    df_ess |> 
        group_by(Round = start_year, agea) |> 
        summarise(mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T)) |> 
        ggplot(aes(x = agea, y = mean)) +
        geom_point(size = 0.5) +
        geom_smooth(method = 'lm') +
        ylim(3,5) +
        facet_wrap(~Round, nrow=2) + 
        labs(
            x = "Age", 
            y = "Government should reduce income differences", 
            caption = "Source: European Social Survey"
        )+
        theme_classic() +
        theme(plot.caption = element_text(hjust = 0))

save_display(fig_age_round, w=8, h=5)
```

There is robust evidence for the association between social class and overall welfare attitudes in Europe [@svallfors_contested_2012; @langsaether_more_2020].The analysis by social class in Ireland suggests a negative relationship between class and the main ESS response variable. The mean values shown in @fig-class5 indicate that 'unskilled workers' show higher levels of agreement with the government reducing income differences whereas 'small business owners' and the 'service class' show lower levels of agreement. 

```{r}
#| label: fig-class5
#| fig-cap: "Scatter plot for 'income differences' by social class (5 groups)"

fig_class5 <- 
    df_ess |> 
        filter(essround == 11 & !is.na(class5)) |> 
        mutate(class = as_factor(class5)) |> 
        group_by(class) |> 
        summarise(
            mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T),
            se = sd(gincdif_inv, na.rm = T)/sqrt(length((gincdif_inv)))
        ) |> 
        ggplot(aes(x = class, y = mean,  ymin = mean - 1.96*se, ymax= mean + 1.96*se)) +
        geom_pointrange() +
        coord_flip() +
        theme_minimal() +
        theme(axis.text.y = element_text(size = 10), plot.caption = element_text(hjust=-0.9)) +
        labs(
            x = "Social Class", 
            y = 'Government should reduce income differences', 
            caption = "Source: European Social Survey"
        )

save_display(fig_class5, w = 6, h=3)

```


## Left-wing voters are more supportive of redistribution

The ESS survey also finds a consistent pattern of association between the respondents' ideological positioning and support for redistribution. Those participants who position themselves more to the left of the left-right political scale tend to be more supportive of redistribution across all rounds of the survey since 2002.

```{r}
#| include: false

df_ess |> 
    filter(!is.na(lrscale) & !is.na(gincdif_inv)) |>
    group_by(essround) |> 
    summarise(cor = cor(lrscale, gincdif_inv))

```

With the data from the Voter Survey it is also possible to examine differences in redistribution preferences across voting choice in the European Parliament elections. People Before Profit (7.5) and Sinn Féin (6.4) voters indicate the highest support, whereas Green Party (4.7) and Fine Gael (4.7) voters reported the lowest support among the main parties. Even when controlling for positioning in the left-right scale, differences between the former two and latter two parties are statistically significant.  These estimates should be interpreted with caution due to the small sample size of respondents who voted in the last European elections. However,  @attewell_deservingness_2021 indicate that this pattern is consistent across most Western European countries, and welfare attitudes are relevant predictors of voting even for parties with focus on non-economic issues such as 'immigration' and EU integration. 

```{r}
#| include: false

df_vs |> 
    filter(study == "EES24") |> 
    group_by(as_factor(party), .drop=TRUE) |> 
        summarise(VS = weighted.mean(redist_inv, w = weight, na.rm = T), n = n())

df_vs$party_lable <-  relevel(as_factor(df_vs$party), ref = "Fine Gael")

lm(redist_inv ~  as_factor(party_lable) + lrscale, data = df_vs |> 
    filter(study == "EES24"), weights = weight) |> summary()



```

## Age, gender and social class are relevant even controlling for occupation and financial situation

We created a fixed-effects multivariate regression model to assess the extent to which these individual factors are associated to support for income redistribution, considering all ESS rounds. In addition to these main variables, the model also includes a measurement of the household financial situation. A subjective measurement was chosen ('Sastisfaction with the household income') instead of household income due to the high proportion of missing values in the former. Finally, the model also includes the following control variables: economic activity ('Main activity in the last seven days'), educational attainment, and political ideology ('Placement in the left-right scale'). 

@tbl-multiv-gincdif below shows the coefficients of the OLS regression model. Despite the overall explanatory power of the model being low (R² = 0.06), the model identifies significant differences across groups.

In terms of age, taking the group of 18 to 24 as reference, all other age groups are expected to have higher support for redistribution, which is the opposite relationship found for the latest round in 2023. In addition, both the social class and subjective income suggest that respondents in more precarious economic situation tend to be more favourable of the government reducing income differences. 

Finally, as discussed earlier in this chapter, both female respondents and those who place themselves on the left of the left-right scale are more likely to support income redistribution, even when controlling for other factors such as age and social class.

```{r}
#| label: tbl-multiv-gincdif
#| tbl-cap: "OLS regression coefficients for 'income differences' "

df_ess$age_group  <- cut(df_ess$agea, c(0, 24, 35, 45, 55, 65, Inf), c("18-24", "25-35", "36-45", "46-55", "56-65", ">66"), include.lowest=TRUE)

df_ess$age_group <-  relevel(df_ess$age_group, ref = "18-24")

df_ess$false_cohort  <-  cut(df_ess$yrbrn, breaks = c(1900, 1940, 1960, 1980, 2020), right = FALSE, labels = c("1900-1940", "1941-1960", "1961-1980", "1981-"))

ind_vars <- c('agea', 'eisced', 'as_factor(gndr)', 'as_factor(class5)', 'as_factor(hincfel)', 'as_factor(mnactic)', 'lrscale', 'as_factor(essround)')

ind_vars_print <- ind_vars[!ind_vars %in% c('eisced', 'as_factor(mnactic)', 'as_factor(essround)')]

tbl_multiv_gincdif <- 
    lm(
        as.formula(paste0("gincdif_inv ~ ", paste(ind_vars, collapse= "+"))), 
        weights = pspwght, 
        data = df_ess 
    )  |> 
    tbl_regression(
        include = ind_vars_print,
        label = list(
            #age_group = "Age group", 
            `as_factor(gndr)` = "Gender", 
            `as_factor(class5)` = 'Social class',
            `as_factor(hincfel)` = 'Financial situation'
        )
    )  |> 
    bold_p(t = 0.05, q = FALSE) |> 
    as_gt() |>
    opt_row_striping()  |> 
    opt_vertical_padding(scale = 0.4) |> 
    tab_style(
        style = cell_fill(color = "#1F355E"),
        locations = cells_column_labels(
        columns = everything()
    )) |> 
    tab_style(
        style = cell_text(color = "white"),
        locations = cells_column_labels(
        columns = everything()
    )) |> 
    tab_footnote("Note: Economic activity, Educational attainment and ESS round were omitted from the table")  |> 
    tab_footnote("Source: European Social Survey")

save_display(tbl_multiv_gincdif)

```


# What people think about welfare policies? {#sec-what} 

The data analysed in this chapter refers to a special module from the European Social Survey on welfare attitudes that was fielded in 2009 (Round 4) and repeated in 2016 (Round 8) (see @meuleman_past_2018 for more on this module). These ronds of data collection contain additional questions on beliefs about welfare, social benefits and their beneficiaries. 


## Meritocratic differences are acceptable for younger and right-wing respondents

As discussed in the previous chapter, a few socio-demographic characteristics such as age and social class may influence people's views on redistribution. The connection between these two factors, social position and redistributive preferences, tend to be explained by self-interest, meaning that people may support redistribution if they believe that it would improve their own material conditions and lives. However, a large body of research shows that other factors related to peoples' views on fairness and meritocracy are also relevant to understand overall support to welfare policies. 

In the 2016 ESS, respondents were asked to what extent they agree with the statement that *‘Large differences in people’s incomes are acceptable to properly reward differences in talents and efforts’*. The majority of participants (`r  df_ess |> filter(!is.na(dfincac) & essround == 8) |> summarise(round(weighted.mean(dfincac %in% c(1,2), w = pspwght)*100))`%) either "agree" or "agree strongly" with the statement.

A multivariate analysis with the same explanatory variables from the model  of the @tbl-multiv-gincdif suggests that gender and class are not statistically significant predictors of agreement. However, older respondents and those who identify more with the political right tend to agree more with this statement.

```{r}
#| include: false
lm(as.formula(paste("dfincac ~ ", paste(ind_vars, collapse= "+"))), 
        weights = pspwght, data = df_ess)  |> 
    tbl_regression() |> 
    bold_p(t = 0.05, q = FALSE)

```

## The government should have more responsibility towards old citizens and working parents compared to the unemployed 

A crucial aspect of the sustainability of modern welfare states is the public support given to its policies. A vast literature in social sciences has addressed the social legitimacy of welfare policies that are targeted at specific groups [@van_oorschot_social_2017], as they create contentious views on which social groups are more or less deserving of the State attention. In the ESS, respondents were asked to indicate how much responsibility governments should have to the following:

"... ensure sufficient childcare services for working parents"   

"... ensure a reasonable standard of living for the old"  

"...ensure a reasonable standard of living for the unemployed"  


In the @fig-deservingness, higher values denote higher government responsibility. On average, respondents from all countries (except The Netherlands) believe that the government should be least responsible for the unemployed compared to working parents and older citizens.   

```{r}
#| label: fig-deservingness
#| fig-cap: "Mean values for deservingness of government support by group and country, 2016"

fig_deservingness <- 
    df_esseu |>
        filter(essround == 8) |> 
        group_by(country = if_else(cntry == 'IE', 'Ireland', cntry))   |> 
        summarise(across(c("gvcldcr", "gvslvol", "gvslvue"), list(
            mean = ~ weighted.mean(.x, w = pspwght, na.rm = T),
            se = ~ sd(.x, na.rm = T)/sqrt(length((.x))))
        ))  |> 
        pivot_longer(cols = -country, names_to = c("Group", ".value"), names_sep = "_") |> 
        mutate(Group = case_match(Group, "gvcldcr" ~ "Working parents", "gvslvol" ~ "Old", "gvslvue" ~ "Unemployed")) |> 
        ggplot(aes(x = reorder(country, -mean), y = mean, ymin = mean - 1.96*se, ymax = mean + 1.96*se,  group = Group, color = Group)) +
        geom_pointrange() +
        coord_flip() +
        theme_classic() +
        theme(
            legend.position="bottom",
            legend.text = element_text(size = 16),
            axis.text.y = element_text(size = 16)
        )

save_display(fig_deservingness, h=7, w=6)

```

In Ireland, a greater prioritisation of the older over the unemployed is found mainly among respondents who are female, in paid work, and with lower level of educational attainment. 

```{r}
#| include: false

# Create variable indicating priority to unemployed over old

df_ess$old_prior <- df_ess$gvslvol - df_ess$gvslvue

lm(as.formula(paste("old_prior ~ ", paste(ind_vars, collapse= "+"), '+ uemplwk')), 
        weights = pspwght, data = df_ess)  |> 
    tbl_regression() |> 
    bold_p(t = 0.05, q = FALSE)

```


## The view that 'Social benefits make people lazy' is more prevalent in Ireland compared to the European average  

The module also included questions in relation to beliefs about the consequences of social benefits and services: 

::: {.callout-note collapse="true"}

### Question wording

Using this card please tell me to what extent you agree or disagree that social benefits and services in [country]... 

"...cost businesses too much in taxes/charges?"

"...lead to a more equal society?"  

"... make people lazy?" 

"...make people less willing care for one another?"

"...prevent widespread poverty"  

"...place too great strain on economy"

:::

The average level of agreement remains similar comparing 2008 and 2016. The only exception is the average for "social benefits place too great strain on economy", which oberved a significant reduction in the level of agreement, as shown in @fig-benefits. 

```{r}
#| label: fig-benefits
#| fig-cap: "Proportion of respondents who agree with statements about social benefits by round"

believe_vars <- c("sbeqsoc", "sblazy", "sblwcoa", "sbprvpv", "sbstrec", "sbbsntx")

# Creating inverted scales so higher values denote higher agreement
df_ess <- df_ess |> mutate(across(all_of(believe_vars), ~ 6 - .x, .names = "{.col}_inv"))
believe_vars_inv <- paste0(believe_vars,"_inv")

# Creating dummy variables for agreement
df_ess <- df_ess |> mutate(across(all_of(believe_vars), ~ if_else(.x %in% c(1,2), 1, 0), .names = "{.col}_agree"))
believe_vars_agree <- paste0(believe_vars,"_agree")

# Creating dummy variables for strong agreement
df_ess <- df_ess |> mutate(across(all_of(believe_vars), ~ if_else(.x == 1, 1, 0), .names = "{.col}_strong_agree"))
believe_vars_strong_agree <- paste0(believe_vars,"_strong_agree")


fig_benefits <- 
    df_ess |>
        filter(essround %in% c(4, 8)) |> 
        group_by(year = as_factor(start_year))  |> 
        summarise(across(all_of(believe_vars_agree), ~ weighted.mean(.x, w = pspwght, na.rm = T))) |> 
        pivot_longer(cols = -year, names_to = "Question", values_to = "mean") |> 
        mutate(Statement = case_when( 
            str_detect(Question, "sbprvpv") ~ "Prevent poverty",
            str_detect(Question, "sblazy") ~ "Make people lazy", 
            str_detect(Question, "sbeqsoc") ~ "Lead to equal society", 
            str_detect(Question, "sbstrec") ~ "Put strain on economy",
            str_detect(Question, "sbbsntx") ~ "Cost businesses too much in taxes",
            str_detect(Question, "sblwcoa") ~ "Make people less willing to care",
        )) |> 
        ggplot(aes(x = year, y = mean, group = Statement, color = Statement)) +
        geom_point() +
        geom_line() +
        scale_color_discrete(breaks=c('Prevent poverty', 'Make people lazy', 'Lead to equal society', 'Put strain on economy', "Cost businesses too much in taxes",'Make people less willing to care')) +
        theme_classic() 

save_display(fig_benefits, w =6, h=3)

```

Despite the reduction from 2009 to 2016, as shown in @fig-lazy, Ireland (58%) had the third highest proportion of respondents agreeing that "social benefits make people lazy", behind Portugal (63%) and Poland (60%). This proportion is significantly lower in countries such as Iceland (24%), Sweden (35%), and Germany (43%).

```{r}
#| label: fig-lazy
#| fig-cap: "Proportion of agreement that social benefits make people lazy, Europe, 2016"

df_esseu <- df_esseu |> mutate(across(all_of(believe_vars), ~ 6 - .x, .names = "{.col}_inv"))
df_esseu <- df_esseu |> mutate(across(all_of(believe_vars), ~ if_else(.x %in% c(1,2), 1, 0), .names = "{.col}_agree"))


tb_lazy <- 
    df_esseu |>
    filter(essround == 8) |> 
    group_by(cntry) |> 
    summarise(
        mean = weighted.mean(sblazy_agree, w = pspwght, na.rm = T))  |> 
    arrange(desc(mean))

fig_lazy <- 
    eumap |> 
        mutate(cntry = case_when(
            name == 'France' ~ "FR",
            name == "Norway" ~ "NO",
            .default = iso_a2)) |> 
        left_join(tb_lazy, by = "cntry") |> 
        ggplot() +
        geom_sf(aes(fill = mean)) +
        scale_fill_distiller(palette = "RdYlBu", na.value = "white") +
        coord_sf(xlim = c(-25, 40), ylim = c(25, 75), expand = FALSE) +
        theme_void() +
        labs(
            fill = "Social benefits make people lazy",
            caption = "Source: European Social Survey, Round 11"
        ) +
        theme(
            legend.position = "left", 
            legend.title = element_text(size = 8),
            plot.caption = element_text(hjust = -1.5)
        )  

save_display(fig_lazy, w =5, h=4)

```



## Compared to male respondents, female respondents are less likely to agree that social benefits leads to equality  

The @tbl-multiv-beliefs below shows logistic regression models for two selected dimensions of beliefs about social benefits. The first model, on social benefits as promoters of equality, no significant differences were found across groups in main variables such as age and social class. However, the model indicates that female respondents are more likely to disagree with the equality outcome compared to male respondents. Another significant difference was found in the 'financial situation' variable, with those who are coping on present income being less likely to agree with the equality statement compared to those living comfortably. Those in worse subjective financial situation ('very difficult on present income') have higher likelihood of agreeing although this difference is not statistically significant. 


```{r}
#| label: tbl-multiv-beliefs
#| tbl-cap: "Logistic regression estimates for agreement on statements about 'social benefits', Ireland, 2016"

model_belief <- function(DV){
    m_form <- as.formula(paste0(DV, " ~ ", paste(ind_vars[!ind_vars %in% c("as_factor(essround)")], collapse= "+")))

    glm(m_form, family = 'binomial', weights = pspwght, data = df_ess  |> filter(essround == 8 )) |> 
        tbl_regression(
            include = ind_vars_print,
            exponentiate = TRUE,
            label = list(`as_factor(gndr)` = "Gender", `as_factor(hincfel)` = "Financial situation")
        )  |> 
        bold_p(t = 0.05, q = FALSE) 
}

tbl_multiv_beliefs <- 
    tbl_merge(
        tbls = map(c("sbeqsoc_agree", "sblazy_agree"), model_belief),
        tab_spanner = c("**Equality**", "**Lazyness**")) |> 
        as_gt() |>
        opt_row_striping()  |> 
        opt_vertical_padding(scale = 0.4) |> 
        tab_style(
            style = cell_fill(color = "#1F355E"),
            locations = cells_column_labels(
            columns = everything()
        )) |> 
        tab_style(
            style = cell_text(color = "white"),
            locations = cells_column_labels(
            columns = everything()
        )) |> 
        tab_footnote("Note: Economic activity, Social class, and ESS round were omitted from the table")  |> 
        tab_footnote("Source: European Social Survey, 2016")

save_display(tbl_multiv_beliefs, vwidth = 800)

```

## Younger, right-wing, and those with lower educational attainment are more likely to agree that social benefits make people lazy

The second model on the @tbl-multiv-beliefs provides a different picture in relation to beliefs on social benefits resulting in lazyness. Respondents in the 18-24 age group are more inclined to agree with this statement compared to  all older age groups. In addition, those who place themselves more to the right of the left-right ideological scale and those with lower educational attainment tend to agree with this statement that social benefits make people lazy. 

## Most respondents indicate that immigrants could receive social benefits after having worked and paid taxes for at least one year

Overall, data from 2016 indicates that about half of respondents in Ireland believe that immigrants should obtain rights to social benefits/services "After worked and paid taxes at least a year". The other options that received most answers were "Once they have become a citizen" (19%) and  "After a year, whether or not have worked" (16%). 

As shown in @fig-immig, only 5% in 2016 indicated that "They should never get the same rights". This proportion is significantly higher in countries such as Hungary (30%) and Czechia (24%). In Ireland, this proportion is also higher for those in worse financial situation (19%) and among those who are unemployed and not looking for job (18%). 


```{r}
#| include: false

df_esseu |>
    filter(essround == 8) |> 
    mutate(never = if_else(imsclbn == 5, 1, 0)) |> 
    group_by(cntry) |> 
    summarise(
        mean = weighted.mean(never, w = pspwght, na.rm = T))  |> 
    arrange(desc(mean)) |> 
    print(n=25)

df_ess |>
    filter(essround == 8) |> 
    mutate(never = if_else(imsclbn == 5, 1, 0)) |> 
    group_by(hincfel) |> 
    summarise(
        mean = weighted.mean(never, w = pspwght, na.rm = T))  |> 
    arrange(desc(mean)) |> 
    print(n=25)


```

```{r}
#| label: fig-immig
#| fig-cap: "Distribution for 'When should immigrants obtain rights to social benefits/services', Ireland, 2016"

fig_immig <- 
    df_ess |>
        filter(essround == 8 & !is.na(imsclbn)) |> 
        count(Moment = as_factor(imsclbn), wt =pspwght) |> 
        mutate(p = round(prop.table(n)*100)) |>  
        ggplot(aes(x = reorder(Moment, p), y = p)) +
        geom_bar(stat = 'identity', width = 0.5, fill = "#223451") +
        coord_flip() +
        theme_classic() +
        theme(axis.text.y = element_text(size = 12))

save_display(fig_immig, w =8, h=4)


```


## Attitudes are impacted by government campaigns

During the 2016 European data collection @power_framing_2022

@dunaiski_does_2025

```{r}
#| include: false

dsp_day <- "2017-04-18"

plot <- 
    df_ess |> 
        filter(!is.na(gincdif_inv)) |> 
        arrange(ess_day) |> 
        group_by(essround) |> 
        mutate(mean = cummean(gincdif_inv)) |> 
        ggplot(aes(x = ess_day, y=mean)) +
        geom_line(stat='identity') +
        # geom_vline(
        #     aes(xintercept = as.Date(dsp_day), yintercept = 0),
        #     color = 'red', linetype = 'dashed'
        # ) +
        #facet_wrap(~essround, scales ='free') +
        ylim(3.7, 3.9) +
        theme_classic()

htmlwidgets::saveWidget(plotly::ggplotly(plot), "temp.html")

```



```{r}

df_ess |> 
    mutate(essround_dsp = case_when(
        essround == 8 & dsp_period == "Before" ~ "8A",
        essround == 8 & dsp_period == "After" ~ "8B",
        .default = as.character(essround)
    )) |> 
    group_by(var_grp = as_factor(age_group), essround_dsp) |> 
    summarise(mean = weighted.mean(gincdif_inv, w= pspwght, na.rm = TRUE)) |> 
    ggplot(aes(x = essround_dsp, y = mean, color = var_grp)) +
    geom_line(aes(group = var_grp))

dsp_models <- map(c("Before", "After"), ~lm(
        as.formula(paste0("gincdif_inv ~ ", paste(ind_vars[ind_vars != "as_factor(essround)"], collapse= "+"))), 
        weights = pspwght, 
        data = subset(df_ess, essround == 8  & abs(dsp_diffdays) < 15 & dsp_period == .x)
    ))  

tbl_merge(map(dsp_models, ~tbl_regression(.x) |> bold_p()))



df_ess |> count(sblazy)

```

## *Redistribution from* and *Redistribution to* 

as pointed out by @cavaille_fair_2025,  

# Has the support changed over time? {#sec-change} 

However, the average support varies substantially over time. The @fig-ess-time below shows the mean values for 'Government should reduce income differences' by round. Overall there is a stable agreement (mean = `r round(weighted.mean(df_ess$gincdif_inv, w = df_ess$pspwght, na.rm = T),1)`). Rounds 5-7 (2010-2014) as well as 10-11(2022-2024) show a slightly higher agreement compared to other rounds and the European average.  

```{r}
#| label: fig-ess-time
#| fig-cap: "Mean values for 'Government should reduce income differences' by round" 

list_countries <- c("IE", "GR", "PT", "GB", "BE", "CZ", "FI", "FR", "DE", "NO", "ES")


df_esseu |>
    filter(cntry %in% list_countries) |> 
    mutate(Country = case_match(cntry,
        "IE" ~ "Ireland",
        "PT" ~ "Portugal",
        # "GR" ~ "Greece",
        "GB" ~ "Great Britain",
        .default = paste0("Other ", length(list_countries), " EU countries"))) |> 
    group_by(round = as_factor(essround), Country)  |> 
    mutate(gincdif_inv = 6 - gincdif) |> 
    summarise(
        mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T),
        se = sd(gincdif_inv, na.rm = T)/sqrt(length((gincdif_inv)))) |> 
    ggplot(aes(x = round, y = mean, group = Country, color = Country)) +
    geom_line()+
    geom_point() +
    ylim(3,5) +
    theme_classic()

    
```

@fig-vs-time shows the density of responses for the scale on support for redistribution of wealth. The mean values are similar for 2014, 2019 and 2024. However, the distribution seems is less skewed in 2024 compared to 2014.

```{r}
#| label: fig-vs-time
#| fig-cap: "Density plot for 'in favour of redistribution of wealth' by survey year"

df_vseu |>
    mutate(
        Country = if_else(as_factor(cntry) == "IRELAND", "Ireland", paste0("Other EU countries")),
        Year = as.numeric(str_remove(study, "EES"))
    ) |> 
    group_by(Year, Country)  |> 
    summarise(
        mean = weighted.mean(redist_inv, w = weight, na.rm = T),
        se = sd(redist_inv, na.rm = T)/sqrt(length((redist_inv)))) |> 
    ggplot(aes(x = Year, y = mean, group = Country, color = Country)) +
    geom_line()+
    geom_point() +
    ylim(4,7) +
    theme_classic()


# df_vseu |> 
#     ggplot(aes(x = redist_inv)) + 
#     geom_density(fill = "lightblue") + 
#     geom_vline(data = df_vs |> group_by(study) |> summarise(mean = weighted.mean(redist_inv, w = weight, na.rm = T)), aes(xintercept = mean), color = 'red') +
#     facet_wrap(~study, nrow = 3) +
#     theme_classic()

```


## Age

The relationship between age and the response variable seems to change across rounds. As shown in @fig-time-age, early rounds suggest that older respondents support more the government reducing income differences, whereas the latest round (11) and round 8 seem to indicate the opposite. 



### ESS

```{r}
#| label: fig-time-age
#| fig-cap: "Scatter plot for 'Government should reduce income differences' by age and round"

fig_ess_age <- 
    df_ess |> 
        group_by(essround, agea) |> 
        summarise(mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T)) |> 
        ggplot(aes(x = agea, y = mean)) +
        geom_point() +
        geom_smooth(method = 'lm') +
        ylim(3,5) +
        facet_wrap(~essround)


save_display(fig_ess_age, w = 7, h =6)

```


### ESS

```{r}
#| label: fig-ess-cohort
#| fig-cap: "Mean values for 'Government should reduce income differences' by false cohort and round"

df_ess |> 
    mutate(false_cohort = cut(yrbrn, breaks = c(1900, 1940, 1980, 2020), right = FALSE, labels = c("1900-1940", "1940-1980", "1980-"))) |>
    filter(!is.na(false_cohort)) |> 
    group_by(essround, false_cohort) |> 
    summarise(mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T)) |> 
    ggplot(aes(x = essround, y = mean, colour = false_cohort, fill = false_cohort)) +
    geom_line()+
    geom_point() +
    ylim(3,5) +
    theme_classic()

```



# Economic crisis and Covid19

As shown in @fig-time-ess, the ESS data indicates a slightly higher average support for redistribution after the 2008 economic crisis and after the 2022 Covid19 pandemic. Evidence from a survey experiment in the USA show that participants are more willing to prioritize society's problems when exposed to issues related to the pandemic [@cappelen_solidarity_2021]. Van Hootegem and Laenen [-@van_hootegem_wave_2023] show that the increase in support for a universal basic income increased with the pandemic but it was short lived. 

In this section we explore further  the potential impact of the pandemic on support for redistribution. 


## Time series

@tbl-multiv-gincdif suggests that both social class and feeling about household income are significant predictors of support for redistribution. @fig-time-hincfel shows that the difference between subjective income groups remains stable over time, with all groups responding similar to 2008 crisis and the pandemic. 

```{r}
#| label: fig-time-hincfel
#| fig-cap: "Average support for redistribution by financial situation (2002-2022)"


get_series <- function(myvar){

df_ess |>
    filter(!is.na(get(myvar))) |> 
    group_by(round = as_factor(essround), class = as_factor(get(myvar)))  |> 
    mutate(gincdif_inv = 6 - gincdif) |> 
    summarise(
        mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T),
        se = sd(gincdif_inv, na.rm = T)/sqrt(length((gincdif_inv)))) |> 
    ggplot(aes(x = round, y = mean, ymin = mean - 1.96*se, ymax = mean + 1.96*se, group = class, color = class)) +
    geom_pointrange() +
    geom_line()+
    ylim(3,4.5)+
    theme_classic()
}

get_series("hincfel")

```

Conversely, the @fig-time-class5 suggests that there is a slightly different pattern in response to the pandemic. There is a trajectory of increasing support for redistribution across skilled and unskilled workers since round 8 (2016). However, for lower-grade service class and small business owners, there seems to be a break in this trajectory captured in the latest wave. 

```{r}
#| label: fig-time-class5
#| fig-cap: "Average support for redistribution by social class (2002-2022)"

get_series("class5")

```


## Job loss


```{r}
#| label: fig-time-jobloss
#| fig-cap: "Average support for redistribution by social class (2002-2022)"

df_ess |> 
    filter(!is.na(class5) & !is.na(hapnoc19)) |> 
    mutate(
        impact = case_when(
            hapljc19 == 1 | hapfoc19 == 1 | hapfuc19 == 1 ~ "Lost job/unpaid leave/Forl.",
            hapirc19 == 1 | hapwrc19 == 1 ~ "Income / hours reduced",
            hapnoc19 == 1 ~ "Nothing",
            is.na(hapnoc19) ~ NA_character_,
            .default = "Other"
        )
    ) |> 
    group_by(impact, class = as_factor(class5)) |> 
    summarise(
        mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T),
        se = sd(gincdif_inv, na.rm = T)/sqrt(length((gincdif_inv)))) |> 
    ggplot(aes(x = class, y = impact, fill = mean))+
    geom_tile() +
    scale_fill_distiller(palette = "RdBu", direction=1) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

    # ggplot(aes(x = class, y = mean, ymin = mean - 1.96*se, ymax = mean + 1.96*se)) +
    # geom_pointrange() +
    # coord_flip()+
    # facet_wrap(~impact, nrow =1) 

```


## Multivariate analysis

We created a multivariate regression model to assess the extent to which these individual factors are associated support for government reducing differences in income levels in Ireland. In order to  The @tbl-multiv-gincdif below shows the coefficients of the OLS regression model. The overall explanatory power of the model is low, indicating that the small set of variables selected do not provide enough information to predict support appropriately. The effect of age seems to be neglectable. Both the social class and subjective income suggest that respondents in more precarious economic situation tend to be more favourable of the government reducing income differences. 

```{r}
##| label: tbl-time-multiv-gincdif
##| tbl-cap: "OLS regression coefficients for 'income differences' "


# lm(gincdif_inv ~ factor(essround) + agea + as_factor(gndr) + as_factor(class5) + as_factor(hincfel) + cso_unemp, weights = pspwght, data = df_ess)  |> 
#     tbl_regression(
#         include = c("agea", "as_factor(gndr)", "as_factor(class5)", "as_factor(hincfel)", "cso_unemp"),
#     )  |> 
#     bold_p(t = 0.05, q = FALSE) |> 
#     add_glance_source_note() |> 
#     as_gt() |>
#     gt::tab_source_note("Time fixed effects (ESS Round) omitted from the table but included in the model") |> 
#     tab_options(table.layout = "auto")


```



# References {.unnumbered}

::: {#refs}
::: 