---
title: "Public support for welfare and redistributive policies in Ireland"
format:
   docx:
    toc: false
    number-sections: true
    link-citations: true
    reference-doc: _assets/esri_template.docx
    fig-cap-location: top
    prefer-html: true
knitr:
  opts_chunk:
    out.width: "70%"
#    pdf:
#     papersize: a4paper
#     documentclass: article
#     link-citations: true
#     fig-cap-location: top
#     fig-pos: "H"
execute: 
  echo: false
  message: false
  warning: false
bibliography: _assets/CFI_Redistribution.bib
---

# Introduction {#sec-introduction} 

Many vulnerable individuals and families in Ireland, each with different profiles, circumstances, and needs, are struggling to make ends meet and are falling into poverty. A significant number of these people rely heavily on welfare benefits, but the current level of support is often insufficient to protect them from poverty. 

In a context of increasing challanges to social welfare in Ireland, it is crucial to have a better understanding of the public support to State actions designed to address these challenges.  

With this objective, the current report presents the findings of a research project that explored attitudes towards welfare and redistributive policies among Irish residents. Ireland offers a particularly relevant context for this research. Over the past few decades, the Irish society experienced different levels of economic growth, income inequality, and poverty @roantree_poverty_2024. In this sense, the changing conditions provide valuable information to examine the factors associated with public support for welfare policies and income redistribution. 

The research project was guide by the following general questions:  
1) **Who** in Ireland supports income redistribution?  

2) **What** are the main attitudes in relation to welfare policies in Ireland?    

3) Has the overall support for welfare policies and income redistribution **changed** over time?   

The [first chapter](@sec-who) in this report addresses the first question of a diffuse and generalised support for income and wealth redistribution in the Irish society. The [second chapter](@sec-what), addressing the second question, examines more specific support to welfare policies, beliefs about social benefits and their benifitiaries. The [final chapter](@sec-change) looks at change over time in relation to these topics and explores the possible impact of the economic recession and the pandemic.

## Key points

- Over time, there is an overall stable support for government's role  in reducing differences in income levels (@fig-ess-time) and for wealth redistribution (@fig-vs-time).

- ESS respondents from lower social classes and in worse financial situation are more likely to agree that the government should reduce differences in income levels (@tbl-multiv-gincdif).  

- From 2008 to 2016, there was a significant reduction on the average agreement that social benefits pplace too great strain on economy (@fig-benefits)  

- Respondents in worse financial situation (subjective self-reported) tend to disagree more with the perspective that social benefits are disincentives for work, economy and social cooperation (@tbl-multiv-beliefs)  


- There is a trajectory of increasing support for redistribution across skilled and unskilled workers since round 8 (2016). However, for lower-grade service class and small business owners, there seems to be a break in this trajectory captured in the latest wave (@fig-class5-time)


## Methods and data

Social researchers make use of a varied set of methods to assess welfare attitudes. However, the main quantitative method in this field is social survey [@steele_attitudes_2019; @svallfors_contested_2012]. Despite their limited capacity to capture nuance and ambivalences in welfare attitudes @goerres_can_2012, surveys help researchers to identify trends and patterns within and across different societies. 

In this sense, the main data source selected for this project is the European Social Survey (ESS)[^ess]. More than 20 European countries participate in this survey that is conducted roughly every two years. Ireland is one of the few countries that participated in all 11 rounds since 2002. On average, 2,200 people who were 18 or older and resident in Ireland were interviewed in each of these rounds. 

This research analysed several questions from a special module on welfare attitudes that was fielded in 2009 and 2016. However, most of the analysis is based on a core question, included in all rounds, on the role of the government in reducing income inequalities. Beyond that, the ESS questionnaire contains several other questions relevant to this research related to socio-demographic characteristics. In addition, based on the class scheme developed by Oesch -@oesch_coming_2006, a "social class" variable was constructed using questions about the respondent's occupation (following @tawfik_script_2020 script).  

[^ess]: For more info, visit https://www.europeansocialsurvey.org 

Finally, this research also relied on public opinion data from Voter Surveys (VS) conducted immediately after the European elections (2014,	2019, 2024). These surveys are part of the European Elections Studies (EES), which also features political elite surveys, media studies, and the Euromanifestos project that codes Party Manifestos. Despite being part of the Eurobarometer project before 1994, thee EES surveys have been conducted independently since 1999. The survey offers additional information particularly on electoral behavious and political attitudes. 



# Who supports redistribution? {#sec-who} 

```{r}
library(tidyverse)
library(haven)
library(gt)
library(gtsummary)
library(rnaturalearth)
library(rnaturalearthdata)

# Function to get the name of the latest file
get_latest <- function(file_pattern){
    file_list <- list.files(path = "./data", pattern = file_pattern)
    latest_file <- tail(sort(file_list), n = 1)
    return(latest_file)
}


# Function to save an image of the table/plot and print the image on the report
save_display  <- function(myobj, w=NULL, h=NULL, vwidth = 600){
    name_str  <- deparse(substitute(myobj))
    file_path  <- paste0("_assets/img/", name_str, '.png')

    if(str_detect(name_str, 'tbl')){
        gtsave(myobj, filename = file_path, vwidth = vwidth)
    } else if (str_detect(name_str, 'fig')) {
        if(missing(w)){
            ggsave(myobj, filename = file_path, dpi=320)
        } else {
            ggsave(myobj, filename = file_path, width = w, height = h, dpi=320)
        }
    } else {
        print("Name of the object should contain 'fig' or 'tbl'")
    }
    
    knitr::include_graphics(file_path)
}

# Read file in the root of sub-folder "data" with name containing "ESS" 
df_ess <- read_dta(paste0("data/", get_latest("ESS")))

# Read file in the root of sub-folder "data" with name containing "VS" 
df_vs <- read_dta(paste0("data/", get_latest("VS_")))

# Read file in the root of sub-folder "data" with name containing "VSEU" 
df_vseu <- read_dta(paste0("data/", get_latest("VSEU_")))

# Read country with cross country data 
df_esseu <- read_dta("./data/ess/ess_eu.dta")

```


The ESS question used to assess support for redistribution asked participants to indicate if the government should take measures to reduce differences in income levels. The interviewer manual does not contain any additional instruction in relation to this question, so respondents provided their answers according to their understanding of the following wording: 

::: {.callout-note}

#### Question wording

Using this card, please say to what extent you agree or disagree with each of the following statements. **The government should take measures to reduce differences in income levels**.   

1 - Agree strongly   

2 - Agree   

3 - Neither agree nor disagree   

4 - Disagree   

5 - Disagree strongly   
:::

The original scale of this variable was inverted for our analysis. So, in the results presented in this report, higher values denote stronger agreement with government reducing income differences.  

Recent studies have raised concerns in relation to the reliability of this question as a measurement of support for redistribution. Margalit and Raviv -@margalit_does_2024 argues that reduction in income differences is too abstract and respondents in general do not link it to redistributive measures. Similarly, it has been suggested that this question only captures a diffuse inclination to equality and that is why responses are inconsistent with voting preferences across countries @dallinger_ambivalence_2022. However, @breznau_moderating_2025 argue that the absence of a relationship between this measurement and other expected correlates (voting preferences, support for concrete policies) is due to the omission of views on government (trust and perceptions of corruption) as a moderating variable. 

Similar to the main ESS question, the VS questionnaires also ask whether voters are in favour of wealth redistribution from the rich to the poor using a scale from 0 ('fully oppose') to 10 ('fully favour'). The original scale of this variable was inverted for this analysis. So, in the results presented here, higher values indicate more in favour of redistribution.

## Europe

The mean levels of support for income redistribution shown in @fig-eumap is in line with the geographical divide identified before @meuleman_past_2018, with @finseraas_income_2009 arguing that countries with higher income inequality tend to be more supportive of income redistribution. The mean value for Ireland (3.95) is slightly higher compared to other countries in the north of Europe such as the United Kingdom (3.82) and the Netherlands (3.63). 

```{r}
#| label: fig-eumap
#| fig-cap: "Mean value for 'Governement should reduce income differences', Europe, 2023"

eumap <- ne_countries(scale = "medium", continent = "Europe", returnclass = "sf") |> filter(!name %in% c("Russia", "Faeroe Is."))

fig_eumap <- 
    eumap |> 
        mutate(cntry = case_when(
            name == 'France' ~ "FR",
            name == "Norway" ~ "NO",
            .default = iso_a2)) |> 
        left_join(
            df_esseu |>
            filter(essround == 11) |> 
            group_by(cntry) |> 
            mutate(gincdif_inv = 6 - gincdif) |> 
            summarise(
                mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T),
                se = sd(gincdif_inv, na.rm = T)/sqrt(length((gincdif_inv)))),
            by = "cntry"
        ) |> 
        ggplot() +
        geom_sf(aes(fill = mean)) +
        scale_fill_distiller(palette = "PuOr", na.value = "grey90") +
        coord_sf(xlim = c(-25, 40), ylim = c(25, 75), expand = FALSE) +
        theme_void() +
        labs(fill = "Government should \nreduce income differences") +
        theme(legend.position = "left", legend.title = element_text(size = 8))  

save_display(fig_eumap, w =5, h=4)
```
Source: European Social Survey


## Gender, age and social class

Despite the profusion of studies on the relationship between gender and welfare policies, there is a gap in understanding the gender dimension of beliefs and perceptions in relation to these policies @garritzmann_gender_2021. Considering the marked gender differences in access to the labour market and levels of pay in Ireland @hingre_gender_2024, it is plausible to also expect differences in demands for income redistirbution. 

However, on average, it is not possible to identify statistically significant differences for male and female individuals in Ireland. The ESS question on the government role has a similar mean values for male (3.96) and female (3.94) respondents. Similarly, the voter study identified similar mean values for wealth redistribution across male (6.3) and female (6.15) participants.

```{r}
#| include: false

# Redistribution
df_ess |> 
    filter(essround == 11) |> 
    group_by(Gender = as_factor(gndr), .drop=TRUE) |>
    summarise(ESS = weighted.mean(gincdif_inv, w = pspwght, na.rm = T))  |> 
    left_join(
        df_vs |> 
        filter(study == "EES24") |> 
        group_by(Gender = as_factor(gndr), .drop=TRUE) |> 
        summarise(VS = weighted.mean(redist_inv, w = weight, na.rm = T))
    ) 
```


On the other hand, there seems to be an age pattern where younger respondents tend to be more supportive of income redistribution in Ireland. As shown in @fig-age-eu, the opposite patter is observed in Germany, Finland, France, and Portugal. Although the UK also has a similar negative relationshop between age and  support, it seems to be less pronounced compared to Ireland. This association is further explored in [Chapter 3](@sec-change) investigating potential generational changes.

```{r}
#| label: fig-age-eu
#| fig-cap: "Scatter plot for 'reduce income differences' by age and country, 2023"

fig_age_eu <- 
    df_esseu |> 
        mutate(gincdif_inv = 6 - gincdif) |> 
        filter(essround == 11 & cntry %in% c('IE', 'FR', 'FI', 'GB', 'NL', 'DE')) |> 
        group_by(Country = as_factor(cntry), agea) |> 
        summarise(mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T)) |> 
        ggplot(aes(x = agea, y = mean)) +
        geom_point() +
        geom_smooth(method = 'lm') +
        ylim(3,5) +
        facet_wrap(~Country) + 
        labs(x = "Age", y = "Government should reduce income differences")+
        theme_classic()

save_display(fig_age_eu, w=5, h=4)
```

The analysis by social class suggests a negative relationship between class and the main ESS response variable. The mean values shown in @fig-class5 indicate that 'unskilled workers' show higher levels of agreement with the government reducing income differences whereas 'small business owners' and the 'service class' show lower levels of agreement. 

```{r}
#| label: fig-class5
#| fig-cap: "Scatter plot for 'income differences' by social class (5 groups)"

fig_class5 <- 
    df_ess |> 
        filter(essround == 11 & !is.na(class5)) |> 
        mutate(class = as_factor(class5)) |> 
        group_by(class) |> 
        summarise(
            mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T),
            se = sd(gincdif_inv, na.rm = T)/sqrt(length((gincdif_inv)))
        ) |> 
        ggplot(aes(x = class, y = mean,  ymin = mean - 1.96*se, ymax= mean + 1.96*se)) +
        geom_pointrange() +
        coord_flip() +
        theme_minimal() +
        theme(axis.text.y = element_text(size = 10)) +
        labs(x = "Social Class", y = 'Government should reduce income differences')

save_display(fig_class5, w = 6, h=3)

```


## Multivariate analysis

We created a multivariate regression model to assess the extent to which these individual factors are associated to support for income redistribution. We also included other variables in the model to control for subjective income ('Sastisfaction with the household income'), economic activity ('Main activity in the last seven days'), and political ideology ('Placement in the left-right scale). 

@tbl-multiv-gincdif below shows the coefficients of the OLS regression model. The overall explanatory power of the model is low (R² = 0.087), indicating that the small set of variables selected do not provide enough information to predict support accurately. However, the model identifies significant differences across groups.

In terms of age, if we take the group of 25 to 34 as reference, all other age groups are expected to have lower support for redistribution. In addition, both the social class and subjective income suggest that respondents in more precarious economic situation tend to be more favourable of the government reducing income differences. 

```{r}
#| label: tbl-multiv-gincdif
#| tbl-cap: "OLS regression coefficients for 'income differences' "

df_ess$age_group  <- cut(df_ess$agea, c(0, 24, 35, 45, 55, 65, Inf), c("18-24", "25-35", "36-45", "46-55", "56-65", ">66"), include.lowest=TRUE)

df_ess$age_group <-  relevel(df_ess$age_group, ref = "25-35")

ind_vars <- c('age_group', 'as_factor(gndr)', 'as_factor(class5)', 'as_factor(hincfel)', 'as_factor(mnactic)', 'lrscale')

tbl_multiv_gincdif <- 
    lm(
        as.formula(paste("gincdif_inv ~ ", paste(ind_vars, collapse= "+"))), 
        weights = pspwght, 
        data = df_ess |> filter(essround == 11) 
    )  |> 
    tbl_regression(
        include = c("age_group", "as_factor(gndr)", "as_factor(class5)"),
    )  |> 
    bold_p(t = 0.05, q = FALSE) |> 
    add_glance_source_note() |> 
    as_gt() |>
    gt::tab_source_note("Note: Omitted control variables: economic activity, subjective income, and political ideology \n Source: European Social Survey, Round 11")  |> 
    opt_row_striping()  |> 
    opt_vertical_padding(scale = 0.4) |> 
    tab_style(
        style = cell_fill(color = "#1F355E"),
        locations = cells_column_labels(
        columns = everything()
    )) |> 
    tab_style(
        style = cell_text(color = "white"),
        locations = cells_column_labels(
        columns = everything()
    ))


save_display(tbl_multiv_gincdif)

```


# What are the main attituddes towards welfare policies? {#sec-what} 

The data analysed in this chapter refers to a special module from the European Social Survey on welfare attitudes that was fielded in 2009 (Round 4) and repeated in 2016 (Round 8) (see @meuleman_past_2018 for more on this module). This module contains additional questions on beliefs about welfare, social benefits and their beneficiaries. 


## Are large differences acceptable?  

As discussed in the previous chapter, a few socio-demographic characteristics such as age and social class may influence people's views on redistribution. In addition to that, other factors related to their view on fairness and meritocracy are also relevant to understand overall support to welfare policies. 

In the 2016 ESS, respondents were asked to what extent they agree with the statement that *‘Large differences in people’s incomes are acceptable to properly reward differences in talents and efforts’*. The majority of participants (`r  df_ess |> filter(!is.na(dfincac) & essround == 8) |> summarise(round(weighted.mean(dfincac %in% c(1,2), w = pspwght)*100))`%) either "agree" or "agree strongly" with the statement.

A multivariate analysis with the same explanatory variables from the model  of the @tbl-multiv-gincdif suggests that gender and class are not statistically significant predictors of agreement. However, older respondents and those who identify more with the political right tend to agree more with this statement.

```{r}
#| include: false
lm(as.formula(paste("dfincac ~ ", paste(ind_vars, collapse= "+"))), 
        weights = pspwght, data = df_ess)  |> 
    tbl_regression() |> 
    bold_p(t = 0.05, q = FALSE)
```


## Who deserves welfare policies?  

A crucial aspect of the sustainability of modern welfare states is the public support given to its policies. A vast literature in social sciences has addressed the social legitimacy of welfare policies that are targeted at specific groups @van_oorschot_social_2017, as they create contentious views on which social groups are more or less deserving of the State attention. In the ESS, respondents were asked to indicate how much responsibility governments should have to the following:

"... ensure sufficient child care services for working parents"   

"... ensure a reasonable standard of living for the old"  

"...ensure a reasonable standard of living for the unemployed"  


In the @fig-deservingness, higher values denote higher government responsibility. On average respondents believe that the government should be more responsible for the old compared to working parents and unemployed. However, there was a reduction in this prioritisation for governments' responsibility in relation to older citizens and unemployed and increase in the average prioritisation for working parents.  

```{r}
#| label: fig-deservingness
#| fig-cap: "Mean values for deservingness of government support by group and round"


fig_deservingness <- 
    df_esseu |>
        filter(essround == 8 & cntry %in% c("IE", "GB", "FR", "PT")) |> 
        group_by(country = as_factor(cntry))   |> 
        summarise(across(c("gvcldcr", "gvslvol", "gvslvue"), list(
            mean = ~ weighted.mean(.x, w = pspwght, na.rm = T),
            se = ~ sd(.x, na.rm = T)/sqrt(length((.x))))
        ))  |> 
        pivot_longer(cols = -country, names_to = c("Group", ".value"), names_sep = "_") |> 
        mutate(Group = case_match(Group, "gvcldcr" ~ "working parents", "gvslvol" ~ "old", "gvslvue" ~ "unemployed")) |> 
        ggplot(aes(x = Group, y = mean, ymin = mean - 1.96*se, ymax = mean + 1.96*se,  group = Group, color = Group)) +
        geom_pointrange() +
        # geom_line() +
        # geom_label(aes(label = country(mean, 1)))+
        facet_wrap(~country) +
        theme_classic() +
        theme(legend.position="bottom")

save_display(fig_deservingness)

```

## What are the consequences of social benefit?  

The module also included questions in relation to beliefs about the consequences of social benefits and services: 

::: {.callout-note collapse="true"}

### Question wording

Using this card please tell me to what extent you agree or disagree that social benefits and services in [country]... 

"...lead to a more equal society?"  

"... make people lazy?" 

"...make people less willing care for one another?"

"...prevent widespread poverty"  

"...place too great strain on economy"

:::

The average level of agreement remains similar comparing 2008 and 2016. The only exception is the average for "social benefits place too great strain on economy", which oberved a significant reduction in the level of agreement, as shown in @fig-benefits. 

```{r}
#| label: fig-benefits
#| fig-cap: "Mean values for beliefs in relation to social benefits by round"

believe_vars <- c("sbeqsoc", "sblazy", "sblwcoa", "sbprvpv", "sbstrec")

df_ess <- df_ess |> mutate(across(all_of(believe_vars), ~ 6 - .x, .names = "{.col}_inv"))

believe_vars_inv <- c("sbeqsoc_inv", "sblazy_inv", "sblwcoa_inv", "sbprvpv_inv", "sbstrec_inv")


df_ess |>
    filter(essround %in% c(4, 8)) |> 
    group_by(round = as_factor(essround))  |> 
    summarise(across(all_of(believe_vars_inv), ~ weighted.mean(.x, w = pspwght, na.rm = T))) |> 
    pivot_longer(cols = -round, names_to = "Question", values_to = "mean") |> 
    mutate(Question = case_match(
        Question, 
        "sbeqsoc_inv" ~ "Lead to equal society", 
        "sblazy_inv" ~ "Make people lazy", 
        "sblwcoa_inv" ~ "Make people less willing to care",
        "sbstrec_inv" ~ "Put strain on economy",
        "sbprvpv_inv" ~ "Prevent poverty"
    )) |> 
    ggplot(aes(x = round, y = mean, group = Question, color = Question)) +
    geom_point() +
    geom_line() +
    # geom_label(aes(label = round(mean, 1)))+
    theme_classic()
    
```


### PCA

A Principal Components Analysis indicate that these five variables can be reduced to the following two latent dimensions named arbitrarily:

- **Protective** (PC1): "Lead to equal society" and "Prevent poverty";    
- **Disincentive** (PC2): "Make people lazy" and "Make people less willing to care", and "Put strain on economy".

```{r}
#| include: false
library(psych)

pca_res <- 
    df_ess |> 
    select(all_of(believe_vars_inv)) |>  
    na.omit() |> 
    psych::principal(nfactors = 2)

pca_vars  <- data.frame(unclass(pca_res$loadings))

```

The variables were aggregated within each dimension using the mean value of the non-missing answers. The @tbl-multiv-beliefs below shows the coefficients of two OLS explanatory models for these two dimensions.

The first model indicates that female respondents tend to provide a slightly lower score on the protective dimension of beliefs about social benefits. None of the other predictors included in the model are statistically significant. 

The second model suggests that the group of "small business owners" and "skilled workers" have higher scores compared to respondents of other social classes. In addition, respondents with worse self-reported financial situation tend to disagree more with beliefs related to disincentive dimension of social beliefs. 

::: {.landscape}


```{r}
#| label: tbl-multiv-beliefs
#| tbl-cap: "OLS regression coefficients for 'beliefs about social benefits' "


vars_protect <- c("sbeqsoc_inv", "sbprvpv_inv")
vars_disinctv <- c("sblazy_inv", "sblwcoa_inv", "sbstrec_inv")

df_ess$protective <- rowMeans(select(df_ess, all_of(vars_protect)), na.rm = T)
df_ess$disincentive <- rowMeans(select(df_ess, all_of(vars_disinctv)), na.rm = T)

model_belief <- function(DV){
    model <- as.formula(paste0(DV, "~ factor(essround) + agea + as_factor(gndr) + as_factor(class5) + as_factor(hincfel)" ))

    lm(model, weights = pspwght, data = df_ess) |> 
        tbl_regression(include = c("agea", "as_factor(gndr)", "as_factor(class5)", "as_factor(hincfel)"))  |> 
        bold_p(t = 0.05, q = FALSE) 
}


tbl_merge(
    tbls = map(c("protective", "disincentive"), model_belief),
    tab_spanner = c("**Protective**", "**Disincentive**")) |> 
    as_gt() |>
    gt::tab_source_note("Time fixed effects (ESS Round) omitted from the table but included in the model") |> 
    tab_options(table.layout = "auto")



```

:::


# Has the support changed over time? {#sec-change} 

However, the average support varies substantially over time. The @fig-ess-time below shows the mean values for 'Government should reduce income differences' by round. Overall there is a stable agreement (mean = `r round(weighted.mean(df_ess$gincdif_inv, w = df_ess$pspwght, na.rm = T),1)`). Rounds 5-7 (2010-2014) as well as 10-11(2022-2024) show a slightly higher agreement compared to other rounds and the European average.  

```{r}
#| label: fig-ess-time
#| fig-cap: "Mean values for 'Government should reduce income differences' by round" 

list_countries <- c("IE", "GR", "PT", "GB", "BE", "CZ", "FI", "FR", "DE", "NO", "ES")


df_esseu |>
    filter(cntry %in% list_countries) |> 
    mutate(Country = case_match(cntry,
        "IE" ~ "Ireland",
        "PT" ~ "Portugal",
        # "GR" ~ "Greece",
        "GB" ~ "Great Britain",
        .default = paste0("Other ", length(list_countries), " EU countries"))) |> 
    group_by(round = as_factor(essround), Country)  |> 
    mutate(gincdif_inv = 6 - gincdif) |> 
    summarise(
        mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T),
        se = sd(gincdif_inv, na.rm = T)/sqrt(length((gincdif_inv)))) |> 
    ggplot(aes(x = round, y = mean, group = Country, color = Country)) +
    geom_line()+
    geom_point() +
    ylim(3,5) +
    theme_classic()

    
```

@fig-vs-time shows the density of responses for the scale on support for redistribution of wealth. The mean values are similar for 2014, 2019 and 2024. However, the distribution seems is less skewed in 2024 compared to 2014.

```{r}
#| label: fig-vs-time
#| fig-cap: "Density plot for 'in favour of redistribution of wealth' by survey year"

df_vseu |>
    mutate(
        Country = if_else(as_factor(cntry) == "IRELAND", "Ireland", paste0("Other EU countries")),
        Year = as.numeric(str_remove(study, "EES"))
    ) |> 
    group_by(Year, Country)  |> 
    summarise(
        mean = weighted.mean(redist_inv, w = weight, na.rm = T),
        se = sd(redist_inv, na.rm = T)/sqrt(length((redist_inv)))) |> 
    ggplot(aes(x = Year, y = mean, group = Country, color = Country)) +
    geom_line()+
    geom_point() +
    ylim(4,7) +
    theme_classic()


# df_vseu |> 
#     ggplot(aes(x = redist_inv)) + 
#     geom_density(fill = "lightblue") + 
#     geom_vline(data = df_vs |> group_by(study) |> summarise(mean = weighted.mean(redist_inv, w = weight, na.rm = T)), aes(xintercept = mean), color = 'red') +
#     facet_wrap(~study, nrow = 3) +
#     theme_classic()

```


## Age

The relationship between age and the response variable seems to change across rounds. As shown in @fig-time-age, early rounds suggest that older respondents support more the government reducing income differences, whereas the latest round (11) and round 8 seem to indicate the opposite. 



### ESS

```{r}
#| label: fig-time-age
#| fig-cap: "Scatter plot for 'Government should reduce income differences' by age and round"

fig_ess_age <- 
    df_ess |> 
        group_by(essround, agea) |> 
        summarise(mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T)) |> 
        ggplot(aes(x = agea, y = mean)) +
        geom_point() +
        geom_smooth(method = 'lm') +
        ylim(3,5) +
        facet_wrap(~essround)


save_display(fig_ess_age, w = 7, h =6)

```


### ESS

```{r}
#| label: fig-ess-cohort
#| fig-cap: "Mean values for 'Government should reduce income differences' by false cohort and round"

df_ess |> 
    mutate(false_cohort = cut(yrbrn, breaks = c(1900, 1940, 1980, 2020), right = FALSE, labels = c("1900-1940", "1940-1980", "1980-"))) |>
    filter(!is.na(false_cohort)) |> 
    group_by(essround, false_cohort) |> 
    summarise(mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T)) |> 
    ggplot(aes(x = essround, y = mean, colour = false_cohort, fill = false_cohort)) +
    geom_line()+
    geom_point() +
    ylim(3,5) +
    theme_classic()

```



# Economic crisis and Covid19

As shown in @fig-time-ess, the ESS data indicates a slightly higher average support for redistribution after the 2008 economic crisis and after the 2022 Covid19 pandemic. Evidence from a survey experiment in the USA show that participants are more willing to prioritize society's problems when exposed to issues related to the pandemic [@cappelen_solidarity_2021]. Van Hootegem and Laenen [-@van_hootegem_wave_2023] show that the increase in support for a universal basic income increased with the pandemic but it was short lived. 

In this section we explore further  the potential impact of the pandemic on support for redistribution. 


## Time series

@tbl-multiv-gincdif suggests that both social class and feeling about household income are significant predictors of support for redistribution. @fig-time-hincfel shows that the difference between subjective income groups remains stable over time, with all groups responding similar to 2008 crisis and the pandemic. 

```{r}
#| label: fig-time-hincfel
#| fig-cap: "Average support for redistribution by financial situation (2002-2022)"


get_series <- function(myvar){

df_ess |>
    filter(!is.na(get(myvar))) |> 
    group_by(round = as_factor(essround), class = as_factor(get(myvar)))  |> 
    mutate(gincdif_inv = 6 - gincdif) |> 
    summarise(
        mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T),
        se = sd(gincdif_inv, na.rm = T)/sqrt(length((gincdif_inv)))) |> 
    ggplot(aes(x = round, y = mean, ymin = mean - 1.96*se, ymax = mean + 1.96*se, group = class, color = class)) +
    geom_pointrange() +
    geom_line()+
    ylim(3,4.5)+
    theme_classic()
}

get_series("hincfel")

```

Conversely, the @fig-time-class5 suggests that there is a slightly different pattern in response to the pandemic. There is a trajectory of increasing support for redistribution across skilled and unskilled workers since round 8 (2016). However, for lower-grade service class and small business owners, there seems to be a break in this trajectory captured in the latest wave. 

```{r}
#| label: fig-time-class5
#| fig-cap: "Average support for redistribution by social class (2002-2022)"

get_series("class5")

```


## Job loss


```{r}
#| label: fig-time-jobloss
#| fig-cap: "Average support for redistribution by social class (2002-2022)"

df_ess |> 
    filter(!is.na(class5) & !is.na(hapnoc19)) |> 
    mutate(
        impact = case_when(
            hapljc19 == 1 | hapfoc19 == 1 | hapfuc19 == 1 ~ "Lost job/unpaid leave/Forl.",
            hapirc19 == 1 | hapwrc19 == 1 ~ "Income / hours reduced",
            hapnoc19 == 1 ~ "Nothing",
            is.na(hapnoc19) ~ NA_character_,
            .default = "Other"
        )
    ) |> 
    group_by(impact, class = as_factor(class5)) |> 
    summarise(
        mean = weighted.mean(gincdif_inv, w = pspwght, na.rm = T),
        se = sd(gincdif_inv, na.rm = T)/sqrt(length((gincdif_inv)))) |> 
    ggplot(aes(x = class, y = impact, fill = mean))+
    geom_tile() +
    scale_fill_distiller(palette = "RdBu", direction=1) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

    # ggplot(aes(x = class, y = mean, ymin = mean - 1.96*se, ymax = mean + 1.96*se)) +
    # geom_pointrange() +
    # coord_flip()+
    # facet_wrap(~impact, nrow =1) 

```


## Multivariate analysis

We created a multivariate regression model to assess the extent to which these individual factors are associated support for government reducing differences in income levels in Ireland. In order to  The @tbl-multiv-gincdif below shows the coefficients of the OLS regression model. The overall explanatory power of the model is low, indicating that the small set of variables selected do not provide enough information to predict support appropriately. The effect of age seems to be neglectable. Both the social class and subjective income suggest that respondents in more precarious economic situation tend to be more favourable of the government reducing income differences. 

```{r}
##| label: tbl-time-multiv-gincdif
##| tbl-cap: "OLS regression coefficients for 'income differences' "


# lm(gincdif_inv ~ factor(essround) + agea + as_factor(gndr) + as_factor(class5) + as_factor(hincfel) + cso_unemp, weights = pspwght, data = df_ess)  |> 
#     tbl_regression(
#         include = c("agea", "as_factor(gndr)", "as_factor(class5)", "as_factor(hincfel)", "cso_unemp"),
#     )  |> 
#     bold_p(t = 0.05, q = FALSE) |> 
#     add_glance_source_note() |> 
#     as_gt() |>
#     gt::tab_source_note("Time fixed effects (ESS Round) omitted from the table but included in the model") |> 
#     tab_options(table.layout = "auto")


```



# References {.unnumbered}

::: {#refs}
::: 